---
title: "Single Nuclear Resolution of ACTA2 Mutation in Thoracic Ascending Aortic Tissue"
author: "Mark E. Pepin, MD, PhD, MS"
date: "02/01/2025"
output:
  html_document: 
    code_folding: hide
    keep_md: true
    toc: true
    toc_float: true
    fig_caption: true
  pdf_document:
    toc: yes
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage[table]{xcolor}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
mainfont: Times
fontsize: 10pt
always_allow_html: yes
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: inline
---

```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
opts_chunk$set(tidy.opts=list(width.cutoff=30),tidy=FALSE, warning = FALSE, message = FALSE, cache = TRUE, cache.lazy = FALSE)
```

**Code Author**: Mark E. Pepin, MD, PhD, MS **Contact**:
[pepinme\@gmail.com](mailto:pepin@broadinstitute.org){.email}\
**Institution**: Brigham and Women's Hospital \| Harvard Medical School \| Stanford Cardiovascular Institute
\| Broad Institute of Harvard and MIT\
**Location**: Stanford, CA, USA

# Environment Setup

```{r setup_folders, eval = FALSE}
# Directories
ifelse(!dir.exists(file.path(paste0("../2_Output/"))), dir.create(file.path(paste0("../2_Output/"))), FALSE)
ifelse(!dir.exists(file.path(paste0("../2_Output/1_ACTA/"))), dir.create(file.path(paste0("../2_Output/1_ACTA/"))), FALSE)
ifelse(!dir.exists(file.path(paste0("../2_Output/1_ACTA/1_QC/"))), dir.create(file.path(paste0("../2_Output/1_ACTA/1_QC/"))), FALSE)
ifelse(!dir.exists(file.path(paste0("../2_Output/1_ACTA/2_Clustering/"))), dir.create(file.path(paste0("../2_Output/1_ACTA/2_Clustering/"))), FALSE)
ifelse(!dir.exists(file.path(paste0("../2_Output/1_ACTA/3_Differential.Expression/"))), dir.create(file.path(paste0("../2_Output/1_ACTA/3_Differential.Expression/"))), FALSE)
ifelse(!dir.exists(file.path(paste0("../2_Output/1_ACTA/4_Trajectory/"))), dir.create(file.path(paste0("../2_Output/1_ACTA/4_Trajectory/"))), FALSE)
ifelse(!dir.exists(file.path(paste0("../2_Output/2_TAA/"))), dir.create(file.path(paste0("../2_Output/2_TAA/"))), FALSE)
ifelse(!dir.exists(file.path(paste0("../2_Output/2_TAA/1_QC/"))), dir.create(file.path(paste0("../2_Output/2_TAA/1_QC/"))), FALSE)
ifelse(!dir.exists(file.path(paste0("../2_Output/2_TAA/2_Clustering/"))), dir.create(file.path(paste0("../2_Output/2_TAA/2_Clustering/"))), FALSE)
ifelse(!dir.exists(file.path(paste0("../2_Output/2_TAA/3_Differential.Expression/"))), dir.create(file.path(paste0("../2_Output/2_TAA/3_Differential.Expression/"))), FALSE)
ifelse(!dir.exists(file.path(paste0("../2_Output/2_TAA/4_Trajectory/"))), dir.create(file.path(paste0("../2_Output/2_TAA/4_Trajectory/"))), FALSE)
ifelse(!dir.exists(file.path(paste0("../2_Output/3_Merged/"))), dir.create(file.path(paste0("../2_Output/3_Merged/"))), FALSE)
ifelse(!dir.exists(file.path(paste0("../2_Output/3_Merged/1_QC/"))), dir.create(file.path(paste0("../2_Output/3_Merged/1_QC/"))), FALSE)
ifelse(!dir.exists(file.path(paste0("../2_Output/3_Merged/2_Clustering/"))), dir.create(file.path(paste0("../2_Output/3_Merged/2_Clustering/"))), FALSE)
ifelse(!dir.exists(file.path(paste0("../2_Output/3_Merged/3_Differential.Expression/"))), dir.create(file.path(paste0("../2_Output/3_Merged/3_Differential.Expression/"))), FALSE)
# Parallelization
library(future)
plan("multicore", workers = 8)
# set this option when analyzing large datasets
options(future.globals.maxSize = 3e+09)
options(Seurat.object.assay.version = "v5")
```

#ACTA2 Mutant Sample \## Uploading 10X Data

Data were analyzed using the Harvard single-cell sequencing core.

```{r Import.Data_acta2, eval = FALSE}
library(dplyr)
library(Seurat)
library(patchwork)
library(openxlsx)
# Load the dataset from the 10X sequenced data
taa.dat <- Read10X(data.dir = "../1_Input/Sequencing/Data_raw")
# Initialize the Seurat object with the raw (non-normalized data).
taa <- CreateSeuratObject(counts = taa.dat, project = "taa_acta2", min.cells = 3, min.features = 200)
taa
```

## Quality Control and Filtering

Standard filtering and QC tehcniques were used to remove poor-quality
reads and nuclei.

```{r QC_acta2, eval = FALSE}
# Create a feature that quantifies the number of mitochondrial counts (impurity)
taa[["percent.mt"]] <- PercentageFeatureSet(taa, pattern = "^MT-")
head(taa@meta.data, 5)
# Visualize QC metrics as a violin plot
VlnPlot(taa, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
# We can visualize feature-feature comparisons using the following 'FeatureScatter'
plot1 <- FeatureScatter(taa, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(taa, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
# The data were filtered to remove nuclei containing >5% mitochondrial counts, as well as unique feature counts greater than 2,500 or less than 200 to minimize the effects of multiplets and empty droplets, respectively.
taa_filtered <- subset(taa, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
taa_filtered
VlnPlot(taa, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0)
```

## Data Normalization and Scaling

A standard global-scaling normalization method "LogNormalize" was used
that normalizes the feature expression measurements for each cell by the
total expression, multiplies this by a scale factor (10,000 by default),
and log-transforms the result.

```{r Normalize_acta2, eval = FALSE}
taa_normalized <- NormalizeData(taa_filtered, normalization.method = "LogNormalize", scale.factor = 10000)
taa_normalized <- FindVariableFeatures(taa_normalized, selection.method = "vst", nfeatures = 10000)
# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(taa_normalized), 10)
# plot variable features with and without labels
plot1 <- VariableFeaturePlot(taa_normalized)
plot2 <- LabelPoints(plot = plot1, points = c(top10,"ACTA2"), repel = TRUE)
plot2
# The next task is to perform linear transformation needed for dimensionality reduction algorithms (PCA, UMAP, etc).
all.genes <- rownames(taa_normalized)
acta2_scaled <- ScaleData(taa_normalized, features = all.genes) # note: the PCA only requires the top X number of most variable genes, as defined by the 'VariableFeatures' command. To make it run faster, we can run the scaling on these alone (which it does by default, i.e. without the 'features' parameter)
```

## Dimensionality Reduction

We can first visualize the primary features/genes driving dimensionality
in our dataset; specifically, the principal components with largest
inter-cellular variability can be seen. Once this is accomplished, we
identify the "dimensionality" of the dataset, or the number of principal
components. From these heuristics, the first 10 dimensions reflect most
of the variabilitiy present in this dataset.

```{r Dimensionality_acta2, eval = FALSE}
# to control for any confounders, we can add the 'vars.to.regress' parameter, including any variable in the metadata <- ScaleData(pbmc, vars.to.regress = "percent.mt")
acta2_scaled <- RunPCA(acta2_scaled, features = VariableFeatures(object = acta2_scaled))
acta2_scaled <- RunTSNE(acta2_scaled, assay = "RNA")
# Visualize the features/genes
VizDimLoadings(acta2_scaled, dims = 1:5, reduction = "pca")
DimPlot(acta2_scaled, reduction = "pca")
DimPlot(acta2_scaled, reduction = "tsne")
pdf(file = "../2_Output/1_ACTA/2_Clustering/PC_Heatmaps.pdf")
DimHeatmap(acta2_scaled, dims = 1:15, cells = 500, balanced = TRUE)
dev.off()
```

## Unsupervised Clustering

Standard methods were used to cluster nuclei based on Uniform Manifold
Approximation and Projection (UMAP), a non-linear dimensionality
reduction technique that projects high-dimensional data into lower
dimensions while preserving its local structure and relationships. It
works by constructing a weighted graph representing the data’s local
neighborhood and then optimizing its layout in a low-dimensional space
to capture both local and global data patterns.

```{r cluster_acta2, eval = FALSE}
# Create a K-nearest neighbor (KNN) graph based on the euclidean distance in PCA space
acta2_scaled <- FindNeighbors(acta2_scaled, dims = 1:10) # uses the number of dimensions identified in the previous step ("dimensionality")
# Now  modularity optimization techniques such as the Louvain algorithm
acta2_scaled <- FindClusters(acta2_scaled, resolution = c(0.3, 0.4, 0.5, 0.6))
# Non-linear dimensional reductino (UMAP/tSNE)
acta2_scaled <- RunUMAP(acta2_scaled, dims = 1:30)
# Create the UMAP:
UMAP_Clustered<-DimPlot(acta2_scaled, reduction = "umap", group.by = "RNA_snn_res.0.3", label = T) + NoLegend()
tSNA_Clustered <- DimPlot(acta2_scaled, reduction = "tsne", group.by = "RNA_snn_res.0.3", label = T) + NoLegend()
UMAP_Clustered+tSNA_Clustered
pdf(file = "../2_Output/1_ACTA/2_Clustering/UMAP_clustered.pdf")
UMAP_Clustered+tSNA_Clustered
dev.off()
Idents(acta2_scaled) <- "RNA_snn_res.0.3"
# Save this instance to avoid needing to re-run:
saveRDS(acta2_scaled, file = "ACTA2_clustered.rds")
```

## Annotated UMAP, Density Plots

Annotation of clusters was accomplished by comparing the expression
patters of canonical cell type marker genes, shown below.

```{r Annotated.Clusters_acta2, eval = FALSE}
library(ggplot2)
library(dplyr)
library(Seurat)
acta2_scaled<-readRDS(file = "ACTA2_clustered.rds")
# Find all gene markers
DEGs_Clusters<-FindAllMarkers(acta2_scaled)
write.csv(DEGs_Clusters, "../2_Output/1_ACTA/DEGs_Clusters.csv")
#Identify Clusters corresponding with known gene markers:
VSMC1_genes<-c("PKD1", "COL6A2", "PDLIM7", "FLNA", "SMTN")
VSMC2_genes<-c("MYH11", "ACTA2", "ITGA8", "PRKG1", "CRISPLD1")
Fibroblast1_genes<-c("ABCA10", "C3", "ADGRD1", "FBLN1", "DCN")
Fibroblast2_genes<-c("UACA", "NFASC", "PRSS23", "SAMD5") # "TEX41",
Fibromyocyte_genes<-c("DGKG", "ADAMTS1", "RGS6", "TNC", "GRIP2") #, "ANGPT2"
EC1_genes<-c("DIPK2B", "ARHGEF15", "STC1", "FLT1") # "NOTCH4",
EC2_genes<-c("VWF", "BMPER", "BMX", "NOS1")
NKT_genes<-c("SKAP1", "RIPOR2", "FYN", "ITGAL", "CD96")
Macrophage_genes<-c("MRC1", "LGMN", "RBPJ", "F13A1", "RBM47")
Dendritic_genes<-c("ITGAX", "S100A9", "CSF3R", "CXCL8")
chondromyocytes <- c("COL2A1", "ACAN", "SOX9")
# Plot density function
library(Nebulosa)
VSMC1_density<-plot_density(acta2_scaled, VSMC1_genes, joint = TRUE, combine = FALSE, pal = "magma")
VSMC2_density<-plot_density(acta2_scaled, VSMC2_genes, joint = TRUE, combine = FALSE, pal = "magma")
Fibroblast1_density<-plot_density(acta2_scaled, Fibroblast1_genes, joint = TRUE, combine = FALSE, pal = "magma")
Fibroblast2_density<-plot_density(acta2_scaled, Fibroblast2_genes, joint = TRUE, combine = FALSE, pal = "magma")
Fibromyocyte_density<-plot_density(acta2_scaled, Fibromyocyte_genes, joint = TRUE, combine = FALSE, pal = "magma")
EC1_density<-plot_density(acta2_scaled, EC1_genes, joint = TRUE, combine = FALSE, pal = "magma")
EC2_density<-plot_density(acta2_scaled, EC2_genes, joint = TRUE, combine = FALSE, pal = "magma")
Macrophage_density<-plot_density(acta2_scaled, Macrophage_genes, joint = TRUE, combine = FALSE, pal = "magma")
NKT_density<-plot_density(acta2_scaled, NKT_genes, joint = TRUE, combine = FALSE, pal = "magma")
Dendritic_density<-plot_density(acta2_scaled, Dendritic_genes, joint = TRUE, combine = FALSE, pal = "magma")
Chondro_density<-plot_density(acta2_scaled, chondromyocytes, joint = TRUE, combine = FALSE, pal = "magma")
Chondro_density
# Print
pdf(file = "../2_Output/1_ACTA/2_Clustering/Density_plots.pdf", height = 3, width = 3)
VSMC1_density
VSMC2_density
Fibroblast1_density
Fibroblast2_density
Fibromyocyte_density
EC1_density
EC2_density
Macrophage_density
NKT_density
Dendritic_density
dev.off()
# Overlay these gene markers onto the UMAP to identify clusters
FeaturePlot(acta2_scaled, reduction = "umap", label = T, features = Fibroblast1_genes)
FeaturePlot(acta2_scaled, reduction = "umap", label = T, features = Fibroblast2_genes)
FeaturePlot(acta2_scaled, reduction = "umap", label = T, features = EC1_genes)
FeaturePlot(acta2_scaled, reduction = "umap", label = T, features = EC2_genes)
FeaturePlot(acta2_scaled, reduction = "umap", label = T, features = NKT_genes)
FeaturePlot(acta2_scaled, reduction = "umap", label = T, features = Macrophage_genes)
FeaturePlot(acta2_scaled, reduction = "umap", label = T, features = VSMC1_genes)
FeaturePlot(acta2_scaled, reduction = "umap", label = T, features = VSMC2_genes)
FeaturePlot(acta2_scaled, reduction = "umap", label = T, features = Dendritic_genes)
#Create a figure of cell-type specific markers overlying the UMAP
pdf(file = "../2_Output/1_ACTA/2_Clustering/Cell.Type_UMAP.pdf", height = 10, width = 15)
FeaturePlot(acta2_scaled, reduction = "umap", label = T, features = c("ACTA2", "FBLN1", "F13A1", "VWF", "STC1", "NFASC", "ITGAL", "ITGAX"))
dev.off()
# Create cluster labels based on the feature plot delineations of cellular phenotypes
acta2_scaled$umap_cluster <- Idents(acta2_scaled) # save old cluster names (for trajectory analysis)
acta2_scaled <- RenameIdents(object = acta2_scaled, # Rename cluster names
                          `0` = "VSMC_1", 
                          `1` = "VSMC_2", 
                          `2` = "VSMC_3",
                          `3` = "Fibroblast",
                          `4` = "Macrophage",
                          `5` = "EC2",
                          `6` = "EC1",
                          `7` = "NKT",
                          `8` = "Dendritic")
acta2_scaled$cell_type <- Idents(acta2_scaled)
acta2_scaled$cell_type <- factor(acta2_scaled$cell_type, 
                     levels = c("VSMC_1",
                                "VSMC_2",
                                "VSMC_3",
                                "Fibroblast", 
                                "EC1",
                                "EC2",
                                "Macrophage",
                                "NKT",
                                "Dendritic"))
acta2_scaled <- SetIdent(acta2_scaled, value = "cell_type")
UMAP_CellTypes<-DimPlot(acta2_scaled, reduction = "umap", label = T, repel = T,label.size = 4,
                        cols = c("coral2", 
                                 "coral4",
                                 "coral3",
                                 "wheat", 
                                 "steelblue4",
                                 "deepskyblue3",
                                 "azure4",
                                 "gray",
                                 "tan2")) + NoLegend()
tSNE_CellTypes<-DimPlot(acta2_scaled, reduction = "tsne", label = T, repel = T,label.size = 4,
                        cols = c("coral2", 
                                 "coral4",
                                 "coral3",
                                 "wheat", 
                                 "steelblue4",
                                 "deepskyblue3",
                                 "azure4",
                                 "gray",
                                 "tan2")) + NoLegend()
UMAP_CellTypes+tSNE_CellTypes
pdf(file = "../2_Output/1_ACTA/2_Clustering/UMAP_Annotated.pdf", height = 5, width = 5)
UMAP_CellTypes
dev.off()
## Use ridgeplots to identify bimodal gene marker distributions (enriched clusters)
pdf(file = "../2_Output/1_ACTA/2_Clustering/Celltype_RidgePlots.pdf", height = 10, width = 15)
RidgePlot(acta2_scaled, features = Fibroblast1_genes, ncol = 2)
RidgePlot(acta2_scaled, features = Fibroblast2_genes, ncol = 2)
RidgePlot(acta2_scaled, features = EC1_genes, ncol = 2)
RidgePlot(acta2_scaled, features = EC2_genes, ncol = 2)
RidgePlot(acta2_scaled, features = NKT_genes, ncol = 2)
RidgePlot(acta2_scaled, features = Macrophage_genes, ncol = 2)
RidgePlot(acta2_scaled, features = VSMC1_genes, ncol = 2)
RidgePlot(acta2_scaled, features = VSMC2_genes, ncol = 2)
RidgePlot(acta2_scaled, features = Dendritic_genes, ncol = 2)
dev.off()
# Differential Expression
# Export DEGs using cell-type clusters
DEGs_CellTypes<-FindAllMarkers(acta2_scaled)
write.csv(DEGs_CellTypes, "../2_Output/1_ACTA/DEGs_Clusters.csv")
#For each cluster
EC1.markers <- FindMarkers(acta2_scaled, ident.1 = "EC1", min.pct = 0.25)
EC2.markers <- FindMarkers(acta2_scaled, ident.1 = "EC2", min.pct = 0.25)
Dendritic.markers <- FindMarkers(acta2_scaled, ident.1 = "Dendritic", min.pct = 0.25)
VSMC1.markers <- FindMarkers(acta2_scaled, ident.1 = "VSMC_1", min.pct = 0.25)
VSMC2.markers <- FindMarkers(acta2_scaled, ident.1 = "VSMC_2", min.pct = 0.25)
Fibroblast.markers <- FindMarkers(acta2_scaled, ident.1 = "Fibroblast", min.pct = 0.25)
Macrophage.markers <- FindMarkers(acta2_scaled, ident.1 = "Macrophage", min.pct = 0.25)
NKT.markers <- FindMarkers(acta2_scaled, ident.1 = "NKT", min.pct = 0.25)
# Create a dot-bplot of the top 5 markers for each 
library(scCustomize)
paletteLength <- 100
myColor <- colorRampPalette(c("dodgerblue4", "white", "coral2"))(paletteLength)
top5_markers <- Extract_Top_Markers(marker_dataframe = DEGs_CellTypes, num_genes = 5, named_vector = FALSE,
    make_unique = TRUE)
pdf(file = "../2_Output/1_ACTA/2_Clustering/Dot.plot_top5markers.pdf", height = 10, width = 7)
Clustered_DotPlot(seurat_object = acta2_scaled, features = top5_markers, k = 10, colors_use_exp = myColor)
dev.off()
# Create excel sheet
library(openxlsx)
wb_DESeq<-createWorkbook()
  addWorksheet(wb_DESeq, "DEGs_all")
  writeData(wb_DESeq, "DEGs_all", DEGs_CellTypes, startCol = 1, rowNames = T)
  addWorksheet(wb_DESeq, "EC1")
  writeData(wb_DESeq, "EC1", EC1.markers, startCol = 1, rowNames = T)
    addWorksheet(wb_DESeq, "EC2")
  writeData(wb_DESeq, "EC2", EC2.markers, startCol = 1, rowNames = T)
    addWorksheet(wb_DESeq, "VSMC1")
  writeData(wb_DESeq, "VSMC1", VSMC1.markers, startCo = 1, rowNames = T)
    addWorksheet(wb_DESeq, "VSMC2")
  writeData(wb_DESeq, "VSMC2", VSMC2.markers, startCo = 1, rowNames = T)
      addWorksheet(wb_DESeq, "Fibroblast")
  writeData(wb_DESeq, "Fibroblast", Fibroblast.markers, startCol = 1, rowNames = T)
    addWorksheet(wb_DESeq, "Macrophage")
  writeData(wb_DESeq, "Macrophage", Macrophage.markers, startCol = 1, rowNames = T)
    addWorksheet(wb_DESeq, "Dendritic")
  writeData(wb_DESeq, "Dendritic", Dendritic.markers, startCol = 1, rowNames = T)
      addWorksheet(wb_DESeq, "NKT")
  writeData(wb_DESeq, "NKT", NKT.markers, startCol = 1, rowNames = T)
saveWorkbook(wb_DESeq, file = "../2_Output/1_ACTA/3_Differential.Expression/DEGs_CellType.Specific.xlsx", overwrite = T)
# Heatmap of Clusters
DEGs_CellTypes %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10
pdf(file = "../2_Output/1_ACTA/2_Clustering/Celltype_Heatmap.pdf", height = 10, width = 15)
DoHeatmap(acta2_scaled, features = top10$gene, size = 3, disp.min = -2, disp.max = 2) + scale_fill_gradientn(colors = c("dodgerblue4", "white", "coral2")) + labs(title = "Heatmap of Top-10 most variable genes within each cluster")
dev.off()
#
# Contractility score
contractility_genes <- c("CNN1", "TAGLN", "ACTA2", "MYH11")
gene_list <- contractility_genes[contractility_genes %in% rownames(acta2_scaled)] 
gene_expr <- GetAssayData(object = acta2_scaled, assay = "RNA", slot = "data")[gene_list, ]
sum_expr <- colSums(gene_expr)
acta2_scaled$contractility <- scale(sum_expr)
#osteogenic score
gene_list <- c("CBFA1", "MSX2", "RUNX2", "SOX9", "SPP1", "BGLAP", "ALPL", "COL2A1")
gene_list <- gene_list[gene_list %in% rownames(acta2_scaled)]
gene_expr <- GetAssayData(object = acta2_scaled, assay = "RNA", slot = "data")[gene_list, ]
sum_expr <- colSums(gene_expr)
acta2_scaled$osteogenic <- scale(sum_expr)
# Synthetic score
gene_list <-c("CCND1","PCNA", "MKI67")
gene_list <- gene_list[gene_list %in% rownames(acta2_scaled)]
gene_expr <- GetAssayData(object = acta2_scaled, assay = "RNA", slot = "data")[gene_list, ]
sum_expr <- colSums(gene_expr)
acta2_scaled$Synthetic <- scale(sum_expr)
# EZH2 Target Score
# Load the package
library(GSEABase)
# Specify the file path to your GMT file
gmt_file <- "../1_Input/EZH2/EZH2.gmt"
# Read the GMT file
gene_sets <- getGmt(gmt_file)
# Explore the data
gene_sets_list <- geneIds(gene_sets) # Extract genes for each set
names(gene_sets_list) # View gene set names
EZH2_Targets <- gene_sets_list[[3]] # View genes in the first set
gene_list <- intersect(EZH2_Targets, rownames(acta2_scaled))
gene_expr <- GetAssayData(object = acta2_scaled, assay = "RNA", slot = "data")[gene_list, ]
sum_expr <- colSums(gene_expr)
acta2_scaled$EZH2_Score <- scale(sum_expr)
# Save file for downstream analysis
saveRDS(acta2_scaled, file = "../1_Input/ACTA2_annotated.rds")
```

## VSMC - Proliferation scores

To explore whether differences in VSMC subpopulations reflects
differential regulation of proliferative pathways, as previously
reported by Kwartler et al. 2023, we used the Gene Ontology term
GO0008283.

```{r proliferation, eval = FALSE}
# Load the necessary libraries
acta2_scaled<-readRDS(file = "../1_Input/ACTA2_annotated.rds")
acta2_scaled <- subset(acta2_scaled, idents = c("VSMC_1", "VSMC_2", "VSMC_3")) # select only the VSMCs
acta2_scaled <- RunPCA(acta2_scaled) # re-cluster
acta2_scaled <- RunUMAP(acta2_scaled, assay = "RNA", dims = 1:40) # re-cluster
Idents(acta2_scaled) <- acta2_scaled$RNA_snn_res.0.3
library(clusterProfiler)
library(org.Mm.eg.db)  # Mouse gene database
# Define the mouse GO term of interest, e.g., GO:0008150 (biological process)
go_term <- "GO:0008283" # used in the original 2023 paper
# Extract gene symbols associated with the specified GO term
genes_vector <- AnnotationDbi::select(
  org.Mm.eg.db,                 # Mouse gene database
  keys = go_term,               # GO term of interest
  columns = "SYMBOL",           # Retrieve the gene symbols
  keytype = "GOALL"             # Define the key type as GO terms
)$SYMBOL
# Proliferation Score
genes_vector <- toupper(genes_vector)
gene_list <- genes_vector[genes_vector %in% rownames(acta2_scaled)] 
gene_expr <- GetAssayData(object = acta2_scaled, slot = "data")[gene_list, ]
sum_expr <- colSums(gene_expr)
acta2_scaled$proliferation <- scale(sum_expr)
pdf(file = "../2_Output/1_ACTA/4_Trajectory/Proliferation.Score_UMAP.pdf", height = 4, width = 4)
FeaturePlot(acta2_scaled, features = "proliferation", reduction = "umap") + ggtitle("Proliferation Score") +
scale_color_gradient2(low = "darkcyan", mid = "white", high = "darkred", midpoint = median(acta2_scaled$proliferation, na.rm = FALSE))
dev.off()
pdf(file = "../2_Output/1_ACTA/4_Trajectory/EZH2.Score_UMAP.pdf", height = 4, width = 4)
FeaturePlot(acta2_scaled, features = "EZH2_Score", reduction = "umap") + ggtitle("EZH2 Target Score") +
scale_color_gradient2(low = "darkcyan", mid = "white", high = "darkred", midpoint = median(acta2_scaled$EZH2_Score, na.rm = FALSE))
dev.off()
# Contractility
go_term <- "GO:0006940" # used in the original 2023 paper
genes_vector <- AnnotationDbi::select(
  org.Mm.eg.db,                 # Mouse gene database
  keys = go_term,               # GO term of interest
  columns = "SYMBOL",           # Retrieve the gene symbols
  keytype = "GOALL"             # Define the key type as GO terms
)$SYMBOL
genes_vector <- toupper(genes_vector)
gene_list <- genes_vector[genes_vector %in% rownames(acta2_scaled)] 
gene_expr <- GetAssayData(object = acta2_scaled, slot = "data")[gene_list, ]
sum_expr <- colSums(gene_expr)
acta2_scaled$contractility <- scale(sum_expr)
pdf(file = "../2_Output/1_ACTA/4_Trajectory/Contractility.Score_UMAP.pdf", height = 4, width = 4)
FeaturePlot(acta2_scaled, features = "contractility", reduction = "umap") + ggtitle("Contractility Score") +
scale_color_gradient2(low = "darkcyan", mid = "white", high = "darkred", midpoint = median(acta2_scaled$contractility, na.rm = FALSE))
dev.off()
```

## VSMC - Trajectory Analysis (Monocle 3)

```{r monocle3_acta2, eval = FALSE}
library(monocle3)
library(Seurat)
library(SeuratWrappers)
library(dplyr)
library(ggtrace)
ann_colors = list(CellType = c(VSMC_1="coral2",
              VSMC_2="coral4",
              VSMC_3="coral3",
                Fibroblast="wheat", 
                EC1="steelblue4",
                EC2="deepskyblue3",
                Macrophage="azure4",
                NKT="gray",
                Dendritic="tan2"))
acta2_scaled<-readRDS(file = "../1_Input/ACTA2_annotated.rds")
# Subsetting and re-normalization
acta2_scaled <- subset(acta2_scaled, idents = c("VSMC_1", "VSMC_2", "VSMC_3")) # select only the VSMCs
acta2_scaled <- RunPCA(acta2_scaled) # re-cluster
acta2_scaled <- RunUMAP(acta2_scaled, assay = "RNA", dims = 1:40) # re-cluster
Idents(acta2_scaled) <- acta2_scaled$RNA_snn_res.0.3
# Clustering
pdf(file = "../2_Output/1_ACTA/4_Trajectory/UMAP_VSMC.pdf", height = 4, width = 4)
DimPlot(acta2_scaled, group.by = "RNA_snn_res.0.3", reduction = "umap", label = T, repel = T,label.size = 4, cols = c("coral2", "coral4", "coral3"))+NoLegend()
dev.off()
# 
DefaultAssay(acta2_scaled) = "RNA" # This changed in seurat V5
# acta2_scaled <- ScaleData(object = acta2_scaled)
cds <- as.cell_data_set(acta2_scaled)
cds <- preprocess_cds(cds, num_dim = 100)
cds <- estimate_size_factors(cds)
# Include gene names (not done by default by the seurat conversion)
rowData(cds)$gene_name <- rownames(cds)
rowData(cds)$gene_short_name <- rowData(cds)$gene_name
#  Run Monocle
cds <- cluster_cells(cds) # 0.000000002 This step creates "partitions" that are used in the trajectory inference
plot_cells(cds, show_trajectory_graph = FALSE, color_cells_by = "partition") # this shows the partitions overlain on the UMAP
cds <- learn_graph(cds, use_partition = TRUE) # creating trajectory inference within each partition
# cds <- order_cells(cds) # Used when setting the nodes; if already known, use the next line
root_cell <- "CAGATTCGTAACGTGA-1" # names(which(cds@principal_graph_aux$UMAP$pseudotime==0))
cds<-order_cells(cds, root_cells = root_cell)
# Plot the pseudotime on UMAP
plot_cells(cds, 
           color_cells_by = "pseudotime", 
           label_branch_points = FALSE, 
           label_leaves = FALSE)
pdf(file = "../2_Output/1_ACTA/4_Trajectory/ACTA2_UMAP_Trajectory_Partition.pdf", height = 3, width = 3)
plot_cells(cds, 
           color_cells_by = "partition", 
           label_branch_points = FALSE, 
           label_leaves = F,
           show_trajectory_graph = F,
           label_roots = F)
dev.off()
pdf(file = "../2_Output/1_ACTA/4_Trajectory/ACTA2_UMAP_Trajectory_Pseudotime.pdf", height = 3, width = 4)
plot_cells(cds, 
           color_cells_by = "pseudotime", 
           label_branch_points = FALSE, 
           label_leaves = F,
           show_trajectory_graph = F,
           label_roots = F)
dev.off()
# Examine specific genes
plot_cells(cds, 
           genes = c("ACTA2"),
           label_cell_groups = FALSE,
           show_trajectory_graph = FALSE,
           min_expr = 3)
# Identify pseudotime
modulated_genes <- graph_test(cds, neighbor_graph="principal_graph", cores=8) # Identify differentially-expressed genes with pseudotime
modulated_genes <- na.omit(modulated_genes) # remove NA's
modulated_genes <- modulated_genes %>% filter(modulated_genes$q_value < 0.05 & modulated_genes$status =="OK") # filter cds results down
modulated_genes <- modulated_genes[order(-modulated_genes$morans_test_statistic), ] # order by moran's test
modulated_genes <- top_n(modulated_genes, 1000, -q_value)
#### Create a heatmap of genes with similar pseudotime kinetics
genes <- row.names(subset(modulated_genes, q_value < 0.05))
openxlsx::write.xlsx(modulated_genes, "../2_Output/1_ACTA2/Trajectory/Pseudotime_DEGs.xlsx")
library(ComplexHeatmap)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(circlize)
library(monocle3)
pt.matrix <- as.data.frame(exprs(cds)[match(genes,rownames(rowData(cds))),order(pseudotime(cds))])
#
cell_names <- colnames(pt.matrix)
Index<-as.data.frame(cds@colData) %>% dplyr::select(cell_type)
Index<-subset(Index, row.names(Index) %in% cell_names)
Index$CellType <- factor(Index$cell_type,
                         levels = c("VSMC_1",
                                    "VSMC_2",
                                    "VSMC_3",
                                    "Fibroblast", 
                                    "EC1",
                                    "EC2",
                                    "Macrophage",
                                    "NKT",
                                    "Dendritic"))
#Can also use "normalized_counts" instead of "exprs" to use various normalization methods, for example:
#normalized_counts(cds, norm_method = "log")
pt.matrix <- t(apply(pt.matrix,1,function(x){smooth.spline(x,df=6)$y})) # Create a spline that smooths the pseudotime-based expression along 6 degrees of freedom.
filtered_matrix <- t(apply(pt.matrix,1,function(x){(x-mean(x))/sd(x)}))
rownames(pt.matrix) <- genes
colnames(pt.matrix) <- cell_names
###########
paletteLength <- 20
myColor <- colorRampPalette(c("dodgerblue4", "white", "goldenrod1"))(paletteLength)
heatmap_DMC<-pheatmap::pheatmap(pt.matrix, scale="row", 
      cluster_cols = F, 
      cluster_rows = TRUE,
      cutree_rows = 4,
      fontsize_col = 8,
      color = myColor,
      annotation_col = Index,
      annotation_colors = ann_colors,
      show_colnames = F,
      show_rownames = F,
      # right_annotation = ha,
      border_color = NA)
################################
hc <-heatmap_DMC$tree_row
lbl <- cutree(hc, 4)
cluster1<-which(lbl==1)
cluster2<-which(lbl==2)
cluster3<-which(lbl==3)
cluster4<-which(lbl==4)
Cluster1_data<-pt.matrix[cluster1,]
Cluster2_data<-pt.matrix[cluster2,]
write.csv(Cluster2_data, "Cluster2.csv")
Cluster3_data<-pt.matrix[cluster3,]
Cluster4_data<-pt.matrix[cluster4,]
Cluster1_GENES <- rownames(Cluster1_data)
Cluster2_GENES <- rownames(Cluster2_data)
Cluster3_GENES <- rownames(Cluster3_data)
Cluster4_GENES <- rownames(Cluster4_data)
##Enrichr
library(enrichR)
library(dplyr)
dbs <- c("WikiPathway_2023_Human")
enriched_path__1 <- enrichr(Cluster1_GENES, dbs)
enrich_path_1<-enriched_path__1[[dbs]] %>% filter(Adjusted.P.value < 0.05)
head(enrich_path_1)
enriched_path__2 <- enrichr(Cluster2_GENES, dbs)
enrich_path_2<-enriched_path__2[[dbs]] %>% filter(Adjusted.P.value < 0.05)
head(enrich_path_2)
enriched_path__3 <- enrichr(Cluster3_GENES, dbs)
enrich_path_3<-enriched_path__3[[dbs]] %>% filter(Adjusted.P.value < 0.05)
head(enrich_path_3)
enriched_path__4 <- enrichr(Cluster4_GENES, dbs)
enrich_path_4<-enriched_path__4[[dbs]] %>% filter(Adjusted.P.value < 0.05)
head(enrich_path_4)
library(openxlsx)
wb_DESeq<-createWorkbook()
  addWorksheet(wb_DESeq, "Cluster 1")
  writeData(wb_DESeq, "Cluster 1", enrich_path_1, startCol = 1)
  addWorksheet(wb_DESeq, "Cluster 2")
  writeData(wb_DESeq, "Cluster 2", enrich_path_2, startCol = 1)
  addWorksheet(wb_DESeq, "Cluster 3")
  writeData(wb_DESeq, "Cluster 3", enrich_path_3, startCol = 1)
  addWorksheet(wb_DESeq, "Cluster 4")
  writeData(wb_DESeq, "Cluster 4", enrich_path_4, startCol = 1)
saveWorkbook(wb_DESeq, file = paste0("../2_Output/1_ACTA/4_Trajectory/Trajectory_Pathway.xlsx"), overwrite = TRUE)
dbs <- c("GWAS_Catalog_2023")
enriched_1 <- enrichr(Cluster1_GENES, dbs)
enrich_1<-enriched_1[[dbs]] %>% filter(Adjusted.P.value < 0.05)
head(enrich_1)
enriched_2 <- enrichr(Cluster2_GENES, dbs)
enrich_2<-enriched_2[[dbs]] %>% filter(Adjusted.P.value < 0.05)
head(enrich_2)
enriched_3 <- enrichr(Cluster3_GENES, dbs)
enrich_3<-enriched_3[[dbs]] %>% filter(Adjusted.P.value < 0.05)
head(enrich_3)
enriched_4 <- enrichr(Cluster4_GENES, dbs)
enrich_4<-enriched_4[[dbs]] %>% filter(Adjusted.P.value < 0.05)
head(enrich_4)
library(openxlsx)
wb_DESeq<-createWorkbook()
  addWorksheet(wb_DESeq, "Cluster 1")
  writeData(wb_DESeq, "Cluster 1", enrich_1, startCol = 1)
  addWorksheet(wb_DESeq, "Cluster 2")
  writeData(wb_DESeq, "Cluster 2", enrich_2, startCol = 1)
  addWorksheet(wb_DESeq, "Cluster 3")
  writeData(wb_DESeq, "Cluster 3", enrich_3, startCol = 1)
  addWorksheet(wb_DESeq, "Cluster 4")
  writeData(wb_DESeq, "Cluster 4", enrich_4, startCol = 1)
saveWorkbook(wb_DESeq, file = paste0("../2_Output/1_ACTA/4_Trajectory/Trajectory_GWAS.Traits.xlsx"), overwrite = TRUE)
library(stringr)
Aortic_genes <- enrich_2 %>% filter(str_detect(Term,"Aort"))
Aortic_genes <- Aortic_genes$Genes
Aortic_genes <- unique(unlist(strsplit(Aortic_genes, ";", fixed = T)))
# proliferative score
gene_list <-Aortic_genes
gene_list <- gene_list[gene_list %in% rownames(acta2_scaled)]
gene_expr <- GetAssayData(object = acta2_scaled, assay = "RNA", slot = "data")[gene_list, ]
sum_expr <- colSums(gene_expr)
acta2_scaled$Aortopathy <- scale(sum_expr)
Cluster1_Names<-unlist(strsplit(enrich_1$Genes[1], ";", fixed = T))
Cluster2_Names<-unlist(strsplit(enrich_2$Genes[1], ";", fixed = T))
Cluster3_Names<-unlist(strsplit(enrich_3$Genes[1], ";", fixed = T))
Cluster4_Names<-unlist(strsplit(enrich_4$Genes[1], ";", fixed = T))
pdf(file = "../2_Output/1_ACTA/4_Trajectory/Aortic_genes.pdf")
plot_cells(cds, 
           genes = Aortic_genes,
           label_cell_groups = F,
           label_roots = F,
           label_branch_points = F,
           label_leaves = F,
           show_trajectory_graph = F,
           min_expr = 1,
           alpha = 0.8,
           trajectory_graph_color = "grey28",
           ) +
  theme(
  axis.text = element_text(size = 6),    # Adjust the size as needed
  axis.title = element_text(size = 8),  # Adjust the size as needed
  legend.text = element_text(size = 4),  # Adjust the size as needed
  legend.title = element_text(size = 6),
  legend.key.size = unit(2, 'mm')) +
  scale_colour_viridis_c(option = "inferno")
dev.off()

TAA.combined <- readRDS(file = "../1_Input/TAA_snRNA.rds")
library(patchwork)
  plots_up <- VlnPlot(TAA.combined, features = Aortic_genes, split.by = "Disease", group.by = "cell_type",
      pt.size = 0, combine = FALSE)
  plots_up <- lapply(seq_along(plots_up), function(i) {
    plot <- plots_up[[i]] + theme(axis.title.x = element_blank())  # Remove x-axis label
    if (i == length(plots_up)) {
      return(plot)  # Keep the legend in the last plot
    } else {
      return(plot + NoLegend())  # Remove legend in other plots
    }
  })
  pdf(paste0("../2_Output/3_Merged/3_Differential.Expression/Conserved.pdf"), height = 6, width = 10)
  print(wrap_plots(plots = plots_up, ncol = 6))
  dev.off()

############################################################################
GENES_HM<-c(Aortic_genes) #
pt.df<-as.data.frame(pt.matrix)
pt.df$gene_name<-rownames(pt.matrix)
ha = rowAnnotation(link = anno_mark(at = which(pt.df$gene_name %in% GENES_HM),
                   labels = as.character(pt.df[which(pt.df$gene_name %in% GENES_HM), "gene_name"]),
                   labels_gp = gpar(fontsize = 8),
                   padding = unit(1, "mm"))
                   )
heatmap_combination<-ComplexHeatmap::pheatmap(pt.matrix, scale="row",
                    cluster_cols = F,
                    cluster_rows = T,
                    cutree_rows = 4,
                    # cutree_cols = 3,
                     fontsize_col = 5,
                     color = myColor,
                    annotation_names_col = FALSE,
                    show_colnames = F,
                     show_rownames = F,
                     border_color = NA,
                    annotation_colors = NA,
                    right_annotation = ha,
                    annotation_col = Index,
                    border = TRUE)
pdf(file = "../2_Output/1_ACTA/4_Trajectory/Trajectory_Heatmap.pdf", height = 6, width = 6)
heatmap_combination
dev.off()
pdf(file = "../2_Output/1_ACTA/4_Trajectory/Aortopathy.Score_UMAP.pdf", height = 4, width = 4)
FeaturePlot(acta2_scaled, features = "Aortopathy", reduction = "umap") + ggtitle("Aortopathy Score") +
scale_color_gradient2(low = "dodgerblue3", mid = "white", high = "goldenrod2", midpoint = median(acta2_scaled$Aortopathy, na.rm = FALSE))
dev.off()
```

# TAA Control FFPE Sample

Data were analyzed using the Harvard single-cell sequencing core.

```{r Import.Data_con, eval=FALSE}
library(dplyr)
library(Seurat)
library(patchwork)
library(openxlsx)
# Load the dataset from the 10X sequenced data
taa.dat <- Read10X(data.dir = "../1_Input/Sequencing/bri2327_TAA/seurat")
# Initialize the Seurat object with the raw (non-normalized data).
taa <- CreateSeuratObject(counts = taa.dat, project = "taa_acta2", min.cells = 3, min.features = 200)
taa
```

## Quality Control and Filtering

Standard filtering and QC tehcniques were used to remove poor-quality
reads and nuclei.

```{r QC_con, eval=FALSE}
# Create a feature that quantifies the number of mitochondrial counts (impurity)
taa[["percent.mt"]] <- PercentageFeatureSet(taa, pattern = "^MT-")
head(taa@meta.data, 5)
# Visualize QC metrics as a violin plot
VlnPlot(taa, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
# We can visualize feature-feature comparisons using the following 'FeatureScatter'
plot1 <- FeatureScatter(taa, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(taa, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
# The data were filtered to remove nuclei containing >5% mitochondrial counts, as well as unique feature counts greater than 2,500 or less than 200 to minimize the effects of multiplets and empty droplets, respectively.
taa_filtered <- subset(taa, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
taa_filtered
VlnPlot(taa, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0)
```

## Data Normalization

A standard global-scaling normalization method "LogNormalize" was used
that normalizes the feature expression measurements for each cell by the
total expression, multiplies this by a scale factor (10,000 by default),
and log-transforms the result.

```{r Normalize_con, eval=FALSE}
taa_normalized <- NormalizeData(taa_filtered, normalization.method = "LogNormalize", scale.factor = 10000)
```

## Cellular Differential Expression

We sought to identify features/genes exhibiting the highest cell-cell
variability.

```{r cell_diff_con, eval=FALSE}
taa_normalized <- FindVariableFeatures(taa_normalized, selection.method = "vst", nfeatures = 10000)
# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(taa_normalized), 10)
# plot variable features with and without labels
plot1 <- VariableFeaturePlot(taa_normalized)
plot2 <- LabelPoints(plot = plot1, points = c(top10,"ACTA2"), repel = TRUE)
plot2
# The next task is to perform linear transformation needed for dimensionality reduction algorithms (PCA, UMAP, etc).
all.genes <- rownames(taa_normalized)
con_scaled <- ScaleData(taa_normalized, features = all.genes) # note: the PCA only requires the top X number of most variable genes, as defined by the 'VariableFeatures' command. To make it run faster, we can run the scaling on these alone (which it does by default, i.e. without the 'features' parameter)
```

## Dimensionality

We can first visualize the primary features/genes driving dimensionality
in our dataset; specifically, the principal components with largest
inter-cellular variability can be seen. Once this is accomplished, we
identify the "dimensionality" of the dataset, or the number of principal
components. From these heuristics, the first 10 dimensions reflect most
of the variabilitiy present in this dataset.

```{r Dimensionality_TAAng, eval=FALSE}
# to TAAng for any confounders, we can add the 'vars.to.regress' parameter, including any variable in the metadata <- ScaleData(pbmc, vars.to.regress = "percent.mt")
TAAng_scaled <- RunPCA(con_scaled, features = VariableFeatures(object = con_scaled))
TAAng_scaled <- RunTSNE(TAAng_scaled, assay = "RNA")
# Visualize the features/genes
VizDimLoadings(TAAng_scaled, dims = 1:5, reduction = "pca")
DimPlot(TAAng_scaled, reduction = "pca")
DimPlot(TAAng_scaled, reduction = "tsne")
pdf(file = "../2_Output/2_TAA/2_Clustering/PC_Heatmaps.pdf")
DimHeatmap(TAAng_scaled, dims = 1:15, cells = 500, balanced = TRUE)
dev.off()
```

## Clustering

```{r cluster_TAAng, eval=FALSE}
# Create a K-nearest neighbor (KNN) graph based on the euclidean distance in PCA space
TAAng_scaled <- FindNeighbors(TAAng_scaled, dims = 1:10) # uses the number of dimensions identified in the previous step ("dimensionality")
# Now  modularity optimization techniques such as the Louvain algorithm
TAAng_scaled <- FindClusters(TAAng_scaled, resolution = c(0.3, 0.4, 0.5, 0.6))
# Non-linear dimensional reductino (UMAP/tSNE)
TAAng_scaled <- RunUMAP(TAAng_scaled, dims = 1:30)
# Create the UMAP:
UMAP_Clustered<-DimPlot(TAAng_scaled, reduction = "umap", group.by = "RNA_snn_res.0.3", label = T) + NoLegend()
tSNA_Clustered <- DimPlot(TAAng_scaled, reduction = "tsne", group.by = "RNA_snn_res.0.3", label = T) + NoLegend()
UMAP_Clustered+tSNA_Clustered
pdf(file = "../2_Output/2_TAA/2_Clustering/UMAP_clustered.pdf")
UMAP_Clustered+tSNA_Clustered
dev.off()
Idents(TAAng_scaled) <- "RNA_snn_res.0.3"
# Save this instance to avoid needing to re-run:
saveRDS(TAAng_scaled, file = "../1_Input/TAA_nongenomic_clustered.rds")
```

## Gene Markers

```{r Annotated.Clusters_TAAng, eval=FALSE}
library(ggplot2)
library(dplyr)
library(Seurat)
TAAng_scaled<-readRDS(file = "../1_Input/TAA_nongenomic_clustered.rds")
# Find all gene markers
DEGs_Clusters<-FindAllMarkers(TAAng_scaled)
write.csv(DEGs_Clusters, "../2_Output/2_TAA/DEGs_Clusters.csv")
#Identify Clusters corresponding with known gene markers:
VSMC1_genes<-c("PKD1", "COL6A2", "PDLIM7", "FLNA", "SMTN")
VSMC2_genes<-c("MYH11", "ACTA2", "ITGA8", "PRKG1", "CRISPLD1")
Fibroblast1_genes<-c("ABCA10", "C3", "ADGRD1", "FBLN1", "DCN")
Fibroblast2_genes<-c("UACA", "NFASC", "PRSS23", "SAMD5") # "TEX41",
Fibromyocyte_genes<-c("DGKG", "ADAMTS1", "RGS6", "TNC", "GRIP2") #, "ANGPT2"
EC1_genes<-c("DIPK2B", "ARHGEF15", "STC1", "FLT1") # "NOTCH4",
EC2_genes<-c("VWF", "BMPER", "BMX", "NOS1")
NKT_genes<-c("SKAP1", "RIPOR2", "FYN", "ITGAL", "CD96")
Macrophage_genes<-c("MRC1", "LGMN", "RBPJ", "F13A1", "RBM47")
Dendritic_genes<-c("ITGAX", "S100A9", "CSF3R", "CXCL8")
# Plot density function
library(Nebulosa)
VSMC1_density<-plot_density(TAAng_scaled, VSMC1_genes, joint = TRUE, combine = FALSE, pal = "magma")
VSMC2_density<-plot_density(TAAng_scaled, VSMC2_genes, joint = TRUE, combine = FALSE, pal = "magma")
Fibroblast1_density<-plot_density(TAAng_scaled, Fibroblast1_genes, joint = TRUE, combine = FALSE, pal = "magma")
Fibroblast2_density<-plot_density(TAAng_scaled, Fibroblast2_genes, joint = TRUE, combine = FALSE, pal = "magma")
Fibromyocyte_density<-plot_density(TAAng_scaled, Fibromyocyte_genes, joint = TRUE, combine = FALSE, pal = "magma")
EC1_density<-plot_density(TAAng_scaled, EC1_genes, joint = TRUE, combine = FALSE, pal = "magma")
EC2_density<-plot_density(TAAng_scaled, EC2_genes, joint = TRUE, combine = FALSE, pal = "magma")
Macrophage_density<-plot_density(TAAng_scaled, Macrophage_genes, joint = TRUE, combine = FALSE, pal = "magma")
NKT_density<-plot_density(TAAng_scaled, NKT_genes, joint = TRUE, combine = FALSE, pal = "magma")
Dendritic_density<-plot_density(TAAng_scaled, Dendritic_genes, joint = TRUE, combine = FALSE, pal = "magma")
# Print
pdf(file = "../2_Output/2_TAA/2_Clustering/Density_plots.pdf")
VSMC1_density
VSMC2_density
Fibroblast1_density
Fibroblast2_density
Fibromyocyte_density
EC1_density
EC2_density
Macrophage_density
NKT_density
Dendritic_density
dev.off()
# Overlay these gene markers onto the UMAP to identify clusters
FeaturePlot(TAAng_scaled, reduction = "umap", label = T, features = Fibroblast1_genes)
FeaturePlot(TAAng_scaled, reduction = "umap", label = T, features = Fibroblast2_genes)
FeaturePlot(TAAng_scaled, reduction = "umap", label = T, features = EC1_genes)
FeaturePlot(TAAng_scaled, reduction = "umap", label = T, features = EC2_genes)
FeaturePlot(TAAng_scaled, reduction = "umap", label = T, features = NKT_genes)
FeaturePlot(TAAng_scaled, reduction = "umap", label = T, features = Macrophage_genes)
FeaturePlot(TAAng_scaled, reduction = "umap", label = T, features = VSMC1_genes)
FeaturePlot(TAAng_scaled, reduction = "umap", label = T, features = VSMC2_genes)
FeaturePlot(TAAng_scaled, reduction = "umap", label = T, features = Dendritic_genes)
#Create a figure of cell-type specific markers overlying the UMAP
pdf(file = "../2_Output/2_TAA/2_Clustering/Cell.Type_UMAP.pdf", height = 10, width = 15)
FeaturePlot(TAAng_scaled, reduction = "umap", label = T, features = c("TAAng", "FBLN1", "F13A1", "VWF", "STC1", "NFASC", "ITGAL", "ITGAX"))
dev.off()
# Create cluster labels based on the feature plot delineations of cellular phenotypes
TAAng_scaled$umap_cluster <- Idents(TAAng_scaled) # save old cluster names (for trajectory analysis)
TAAng_scaled <- RenameIdents(object = TAAng_scaled, # Rename cluster names
                          `0` = "VSMC", 
                          `1` = "VSMC", 
                          `2` = "Macrophage",
                          `3` = "Macrophage",
                          `4` = "Fibromyocyte",
                          `5` = "Fibroblast",
                          `6` = "NKT",
                          `7` = "EC",
                          `8` = "Dendritic",
                          `9` = "Lymphoid",
                          `10` = "10")
TAAng_scaled$cell_type <- Idents(TAAng_scaled)
TAAng_scaled$cell_type <- factor(TAAng_scaled$cell_type, levels = c("VSMC", 
                                                                    "Fibroblast", 
                                                                    "EC",
                                                                    "Macrophage",
                                                                    "NKT",
                                                                    "Dendritic",
                                                                    "Fibromyocyte",
                                                                    "Lymphoid",
                                                                    "10"))
TAAng_scaled <- SetIdent(TAAng_scaled, value = "cell_type")
UMAP_CellTypes<-DimPlot(TAAng_scaled, reduction = "umap", label = T, repel = T,label.size = 4,
                        cols = c("coral2", 
                                 "wheat", 
                                 "steelblue4",
                                 "azure4",
                                 "gray",
                                 "tan2",
                                 "darkcyan",
                                 "darkgray",
                                 "black")) + NoLegend()
tSNE_CellTypes<-DimPlot(TAAng_scaled, reduction = "tsne", label = T, repel = T,label.size = 4,
                        cols = c("coral2", 
                                 "wheat", 
                                 "steelblue4",
                                 "azure4",
                                 "gray",
                                 "tan2",
                                 "darkcyan",
                                 "darkgray",
                                 "black")) + NoLegend()

UMAP_CellTypes+tSNE_CellTypes
pdf(file = "../2_Output/2_TAA/2_Clustering/UMAP_Annotated.pdf")
UMAP_CellTypes
dev.off()
## Use ridgeplots to identify bimodal gene marker distributions (enriched clusters)
pdf(file = "../2_Output/2_TAA/2_Clustering/Celltype_RidgePlots.pdf", height = 10, width = 15)
RidgePlot(TAAng_scaled, features = Fibroblast1_genes, ncol = 2)
RidgePlot(TAAng_scaled, features = Fibroblast2_genes, ncol = 2)
RidgePlot(TAAng_scaled, features = EC1_genes, ncol = 2)
RidgePlot(TAAng_scaled, features = EC2_genes, ncol = 2)
RidgePlot(TAAng_scaled, features = NKT_genes, ncol = 2)
RidgePlot(TAAng_scaled, features = Macrophage_genes, ncol = 2)
RidgePlot(TAAng_scaled, features = VSMC1_genes, ncol = 2)
RidgePlot(TAAng_scaled, features = VSMC2_genes, ncol = 2)
RidgePlot(TAAng_scaled, features = Dendritic_genes, ncol = 2)
dev.off()
# Differential Expression
# Export DEGs using cell-type clusters
DEGs_CellTypes<-FindAllMarkers(TAAng_scaled)
write.csv(DEGs_CellTypes, "../2_Output/2_TAA/DEGs_Clusters.csv")
#For each cluster
VSMC.markers <- FindMarkers(TAAng_scaled, ident.1 = "VSMC", min.pct = 0.25)
EC.markers <- FindMarkers(TAAng_scaled, ident.1 = "EC", min.pct = 0.25)
Dendritic.markers <- FindMarkers(TAAng_scaled, ident.1 = "Dendritic", min.pct = 0.25)
VSMC.markers <- FindMarkers(TAAng_scaled, ident.1 = "VSMC", min.pct = 0.25)
Fibroblast.markers <- FindMarkers(TAAng_scaled, ident.1 = "Fibroblast", min.pct = 0.25)
Macrophage.markers <- FindMarkers(TAAng_scaled, ident.1 = "Macrophage", min.pct = 0.25)
NKT.markers <- FindMarkers(TAAng_scaled, ident.1 = "NKT", min.pct = 0.25)
# Create a dot-bplot of the top 5 markers for each 
library(scCustomize)
paletteLength <- 100
myColor <- colorRampPalette(c("dodgerblue4", "white", "coral2"))(paletteLength)
top5_markers <- Extract_Top_Markers(marker_dataframe = DEGs_CellTypes, num_genes = 5, named_vector = FALSE,
    make_unique = TRUE)
pdf(file = "../2_Output/2_TAA/2_Clustering/Dot.plot_top5markers.pdf", height = 10, width = 7)
Clustered_DotPlot(seurat_object = TAAng_scaled, features = top5_markers, k = 10, colors_use_exp = myColor)
dev.off()
# Create excel sheet
library(openxlsx)
wb_DESeq<-createWorkbook()
  addWorksheet(wb_DESeq, "DEGs_all")
  writeData(wb_DESeq, "DEGs_all", DEGs_CellTypes, startCol = 1, rowNames = T)
  addWorksheet(wb_DESeq, "EC1")
  writeData(wb_DESeq, "EC1", EC1.markers, startCol = 1, rowNames = T)
    addWorksheet(wb_DESeq, "EC2")
  writeData(wb_DESeq, "EC2", EC2.markers, startCol = 1, rowNames = T)
    addWorksheet(wb_DESeq, "VSMC")
  writeData(wb_DESeq, "VSMC", VSMC.markers, startCo = 1, rowNames = T)
    addWorksheet(wb_DESeq, "Fibroblast")
  writeData(wb_DESeq, "Fibroblast", Fibroblast.markers, startCol = 1, rowNames = T)
    addWorksheet(wb_DESeq, "Macrophage")
  writeData(wb_DESeq, "Macrophage", Macrophage.markers, startCol = 1, rowNames = T)
    addWorksheet(wb_DESeq, "Dendritic")
  writeData(wb_DESeq, "Dendritic", Dendritic.markers, startCol = 1, rowNames = T)
      addWorksheet(wb_DESeq, "NKT")
  writeData(wb_DESeq, "NKT", NKT.markers, startCol = 1, rowNames = T)
saveWorkbook(wb_DESeq, file = "../2_Output/2_TAA/3_Differential.Expression/DEGs_CellType.Specific.xlsx", overwrite = T)
# Heatmap of Clusters
DEGs_CellTypes %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10
pdf(file = "../2_Output/2_TAA/2_Clustering/Celltype_Heatmap.pdf", height = 10, width = 15)
DoHeatmap(TAAng_scaled, features = top10$gene, size = 3, disp.min = -2, disp.max = 2) + scale_fill_gradientn(colors = c("dodgerblue4", "white", "coral2")) + labs(title = "Heatmap of Top-10 most variable genes within each cluster")
dev.off()
#
# Contractility score
contractility_genes <- c("CNN1", "TAGLN", "TAAng", "MYH11")
gene_list <- contractility_genes[contractility_genes %in% rownames(TAAng_scaled)] 
gene_expr <- GetAssayData(object = TAAng_scaled, assay = "RNA", slot = "data")[gene_list, ]
sum_expr <- colSums(gene_expr)
TAAng_scaled$contractility <- scale(sum_expr)
#osteogenic score
gene_list <- c("CBFA1", "MSX2", "RUNX2", "SOX9", "SPP1", "BGLAP", "ALPL", "COL2A1")
gene_list <- gene_list[gene_list %in% rownames(TAAng_scaled)]
gene_expr <- GetAssayData(object = TAAng_scaled, assay = "RNA", slot = "data")[gene_list, ]
sum_expr <- colSums(gene_expr)
TAAng_scaled$osteogenic <- scale(sum_expr)
# Synthetic score
gene_list <-c("CCND1","PCNA", "MKI67")
gene_list <- gene_list[gene_list %in% rownames(TAAng_scaled)]
gene_expr <- GetAssayData(object = TAAng_scaled, assay = "RNA", slot = "data")[gene_list, ]
sum_expr <- colSums(gene_expr)
TAAng_scaled$Synthetic <- scale(sum_expr)
# Save file for downstream analysis
saveRDS(TAAng_scaled, file = "../1_Input/TAAng_annotated.rds")
```

## VSMC-specific Trajectory Analysis (Monocle 3)

```{r monocle3_TAAng, eval=FALSE}
library(monocle3)
library(Seurat)
library(SeuratWrappers)
library(dplyr)
library(ggtrace)
ann_colors = list(CellType = c(VSMC="coral2", 
                             Fibroblast="wheat", 
                             EC="steelblue4",
                             Macrophage="azure4",
                             NKT="gray",
                             Dendritic="tan2",
                             Fibromyocyte="darkcyan",
                             Lymphoid="darkgray",
                             `10`="black"))
TAAng_scaled<-readRDS(file = "../1_Input/TAAng_annotated.rds")
# Subsetting and re-normalization
TAAng_scaled <- subset(TAAng_scaled, idents = "VSMC") # select only the VSMCs
TAAng_scaled <- RunPCA(TAAng_scaled) # re-cluster
TAAng_scaled <- RunUMAP(TAAng_scaled, assay = "RNA", dims = 1:40) # re-cluster
Idents(TAAng_scaled) <- TAAng_scaled$RNA_snn_res.0.3
# Clustering
pdf(file = "../2_Output/2_TAA/4_Trajectory/UMAP_VSMC.pdf", height = 4, width = 4)
DimPlot(TAAng_scaled, group.by = "RNA_snn_res.0.3", reduction = "umap", label = T, repel = T,label.size = 4, cols = c("coral2", "coral4", "coral3"))+NoLegend()
dev.off()
#Contractility
library(ggplot2)
pdf(file = "../2_Output/2_TAA/4_Trajectory/Contractility.Score_UMAP.pdf", height = 4, width = 4)
FeaturePlot(TAAng_scaled, features = "contractility", reduction = "umap") + ggtitle("Contractility Score") +
scale_color_gradient2(low = "darkcyan", mid = "white", high = "darkred", midpoint = median(TAAng_scaled$contractility, na.rm = FALSE))
dev.off()
#Osteogenic
pdf(file = "../2_Output/2_TAA/4_Trajectory/Osteogenicity.Score_UMAP.pdf", height = 4, width = 4)
FeaturePlot(TAAng_scaled, features = "osteogenic", reduction = "umap") + ggtitle("Osteogenicity Score") +
scale_colour_gradient2(low = "darkcyan", mid = "white", high = "darkred", midpoint = median(TAAng_scaled$osteogenic, na.rm = FALSE))
dev.off()
#Synthetic
pdf(file = "../2_Output/2_TAA/4_Trajectory/Synthetic.Score_UMAP.pdf", height = 4, width = 4)
FeaturePlot(TAAng_scaled, features = "Synthetic", reduction = "umap", alpha = 0.6) + ggtitle("Synthetic Score") +
scale_colour_gradient2(low = "darkcyan", mid = "white", high = "darkred", midpoint = mean(TAAng_scaled$Synthetic, na.rm = FALSE))
dev.off()
# 
DefaultAssay(TAAng_scaled) = "RNA" # This changed in seurat V5
# TAAng_scaled <- ScaleData(object = TAAng_scaled)
cds <- as.cell_data_set(TAAng_scaled)
cds <- preprocess_cds(cds, num_dim = 100)
cds <- estimate_size_factors(cds)
# Include gene names (not done by default by the seurat conversion)
rowData(cds)$gene_name <- rownames(cds)
rowData(cds)$gene_short_name <- rowData(cds)$gene_name
#  Run Monocle
cds <- cluster_cells(cds) # 0.000000002 This step creates "partitions" that are used in the trajectory inference
plot_cells(cds, show_trajectory_graph = FALSE, color_cells_by = "partition") # this shows the partitions overlain on the UMAP
cds <- learn_graph(cds, use_partition = TRUE) # creating trajectory inference within each partition
cds <- order_cells(cds) # Used when setting the nodes; if already known, use the next line
root_cell <- "TCATCAGTCAAGCTCC-1" # names(which(cds@principal_graph_aux$UMAP$pseudotime==0))
cds<-order_cells(cds, root_cells = root_cell)
# Plot the pseudotime on UMAP
plot_cells(cds, 
           color_cells_by = "pseudotime", 
           label_branch_points = FALSE, 
           label_leaves = FALSE)
pdf(file = "../2_Output/2_TAA/4_Trajectory/TAAng_UMAP_Trajectory_Partition.pdf", height = 3, width = 3)
plot_cells(cds, 
           color_cells_by = "partition", 
           label_branch_points = FALSE, 
           label_leaves = F,
           show_trajectory_graph = F,
           label_roots = F)
dev.off()
pdf(file = "../2_Output/2_TAA/4_Trajectory/TAAng_UMAP_Trajectory_Pseudotime.pdf", height = 3, width = 4)
plot_cells(cds, 
           color_cells_by = "pseudotime", 
           label_branch_points = FALSE, 
           label_leaves = F,
           show_trajectory_graph = F,
           label_roots = F)
dev.off()
# Examine specific genes
plot_cells(cds, 
           genes = c("ACTA2"),
           label_cell_groups = FALSE,
           show_trajectory_graph = FALSE,
           min_expr = 3)
# Identify pseudotime
modulated_genes <- graph_test(cds, neighbor_graph="principal_graph", cores=8) # Identify differentially-expressed genes with pseudotime
modulated_genes <- na.omit(modulated_genes) # remove NA's
modulated_genes <- modulated_genes %>% filter(modulated_genes$q_value < 0.05 & modulated_genes$status =="OK") # filter cds results down
modulated_genes <- modulated_genes[order(-modulated_genes$morans_test_statistic), ] # order by moran's test
modulated_genes <- top_n(modulated_genes, 1000, -q_value)
#### Create a heatmap of genes with similar pseudotime kinetics
genes <- row.names(subset(modulated_genes, q_value < 0.05))
openxlsx::write.xlsx(modulated_genes, "../2_Output/1_TAAng/Trajectory/Pseudotime_DEGs.xlsx")
library(ComplexHeatmap)
library(ggplot2)
library(dplyr)
library(RColorBrewer)
library(circlize)
library(monocle3)
pt.matrix <- as.data.frame(exprs(cds)[match(genes,rownames(rowData(cds))),order(pseudotime(cds))])
#
cell_names <- colnames(pt.matrix)
Index<-as.data.frame(cds@colData) %>% dplyr::select(cell_type)
Index<-subset(Index, row.names(Index) %in% cell_names)
Index$CellType <- factor(Index$cell_type,
                         levels = c("VSMC", 
                                    "Fibroblast", 
                                    "EC1",
                                    "EC2",
                                    "Macrophage",
                                    "NKT",
                                    "Dendritic"))
#Can also use "normalized_counts" instead of "exprs" to use various normalization methods, for example:
#normalized_counts(cds, norm_method = "log")
pt.matrix <- t(apply(pt.matrix,1,function(x){smooth.spline(x,df=6)$y})) # Create a spline that smooths the pseudotime-based expression along 6 degrees of freedom.
filtered_matrix <- t(apply(pt.matrix,1,function(x){(x-mean(x))/sd(x)}))
rownames(pt.matrix) <- genes
colnames(pt.matrix) <- cell_names
###########
paletteLength <- 20
myColor <- colorRampPalette(c("dodgerblue4", "white", "goldenrod1"))(paletteLength)
heatmap_DMC<-pheatmap::pheatmap(pt.matrix, scale="row", 
      cluster_cols = F, 
      cluster_rows = TRUE,
      cutree_rows = 4,
      fontsize_col = 8,
      color = myColor,
      annotation_col = Index,
      annotation_colors = ann_colors,
      show_colnames = F,
      show_rownames = F,
      # right_annotation = ha,
      border_color = NA)
################################
hc <-heatmap_DMC$tree_row
lbl <- cutree(hc, 4)
cluster1<-which(lbl==1)
cluster2<-which(lbl==2)
cluster3<-which(lbl==3)
cluster4<-which(lbl==4)
Cluster1_data<-pt.matrix[cluster1,]
Cluster2_data<-pt.matrix[cluster2,]
write.csv(Cluster2_data, "Cluster2.csv")
Cluster3_data<-pt.matrix[cluster3,]
Cluster4_data<-pt.matrix[cluster4,]
Cluster1_GENES <- rownames(Cluster1_data)
Cluster2_GENES <- rownames(Cluster2_data)
Cluster3_GENES <- rownames(Cluster3_data)
Cluster4_GENES <- rownames(Cluster4_data)
##Enrichr
library(enrichR)
library(dplyr)
dbs <- c("WikiPathway_2023_Human")
enriched_path__1 <- enrichr(Cluster1_GENES, dbs)
enrich_path_1<-enriched_path__1[[dbs]] %>% filter(Adjusted.P.value < 0.05)
head(enrich_path_1)
enriched_path__2 <- enrichr(Cluster2_GENES, dbs)
enrich_path_2<-enriched_path__2[[dbs]] %>% filter(Adjusted.P.value < 0.05)
head(enrich_path_2)
enriched_path__3 <- enrichr(Cluster3_GENES, dbs)
enrich_path_3<-enriched_path__3[[dbs]] %>% filter(Adjusted.P.value < 0.05)
head(enrich_path_3)
enriched_path__4 <- enrichr(Cluster4_GENES, dbs)
enrich_path_4<-enriched_path__4[[dbs]] %>% filter(Adjusted.P.value < 0.05)
head(enrich_path_4)
library(openxlsx)
wb_DESeq<-createWorkbook()
  addWorksheet(wb_DESeq, "Cluster 1")
  writeData(wb_DESeq, "Cluster 1", enrich_path_1, startCol = 1)
  addWorksheet(wb_DESeq, "Cluster 2")
  writeData(wb_DESeq, "Cluster 2", enrich_path_2, startCol = 1)
  addWorksheet(wb_DESeq, "Cluster 3")
  writeData(wb_DESeq, "Cluster 3", enrich_path_3, startCol = 1)
  addWorksheet(wb_DESeq, "Cluster 4")
  writeData(wb_DESeq, "Cluster 4", enrich_path_4, startCol = 1)
saveWorkbook(wb_DESeq, file = paste0("../2_Output/2_TAA/4_Trajectory/Trajectory_Pathway.xlsx"), overwrite = TRUE)
dbs <- c("GWAS_Catalog_2023")
enriched_1 <- enrichr(Cluster1_GENES, dbs)
enrich_1<-enriched_1[[dbs]] %>% filter(Adjusted.P.value < 0.05)
head(enrich_1)
enriched_2 <- enrichr(Cluster2_GENES, dbs)
enrich_2<-enriched_2[[dbs]] %>% filter(Adjusted.P.value < 0.05)
head(enrich_2)
enriched_3 <- enrichr(Cluster3_GENES, dbs)
enrich_3<-enriched_3[[dbs]] %>% filter(Adjusted.P.value < 0.05)
head(enrich_3)
enriched_4 <- enrichr(Cluster4_GENES, dbs)
enrich_4<-enriched_4[[dbs]] %>% filter(Adjusted.P.value < 0.05)
head(enrich_4)
library(openxlsx)
wb_DESeq<-createWorkbook()
  addWorksheet(wb_DESeq, "Cluster 1")
  writeData(wb_DESeq, "Cluster 1", enrich_1, startCol = 1)
  addWorksheet(wb_DESeq, "Cluster 2")
  writeData(wb_DESeq, "Cluster 2", enrich_2, startCol = 1)
  addWorksheet(wb_DESeq, "Cluster 3")
  writeData(wb_DESeq, "Cluster 3", enrich_3, startCol = 1)
  addWorksheet(wb_DESeq, "Cluster 4")
  writeData(wb_DESeq, "Cluster 4", enrich_4, startCol = 1)
saveWorkbook(wb_DESeq, file = paste0("../2_Output/2_TAA/4_Trajectory/Trajectory_GWAS.Traits.xlsx"), overwrite = TRUE)
library(stringr)
Aortic_genes <- enrich_1 %>% filter(str_detect(Term,"Aort"))
Aortic_genes <- Aortic_genes$Genes
Aortic_genes <- unique(unlist(strsplit(Aortic_genes, ";", fixed = T)))
# proliferative score
gene_list <-Aortic_genes
gene_list <- gene_list[gene_list %in% rownames(TAAng_scaled)]
gene_expr <- GetAssayData(object = TAAng_scaled, assay = "RNA", slot = "data")[gene_list, ]
sum_expr <- colSums(gene_expr)
TAAng_scaled$Aortic <- scale(sum_expr)
Cluster1_Names<-unlist(strsplit(enrich_1$Genes[1], ";", fixed = T))
Cluster2_Names<-unlist(strsplit(enrich_2$Genes[1], ";", fixed = T))
Cluster3_Names<-unlist(strsplit(enrich_3$Genes[1], ";", fixed = T))
Cluster4_Names<-unlist(strsplit(enrich_4$Genes[1], ";", fixed = T))
pdf(file = "../2_Output/2_TAA/4_Trajectory/Aortic_genes.pdf")
plot_cells(cds, 
           genes = Aortic_genes,
           label_cell_groups = F,
           label_roots = F,
           label_branch_points = F,
           label_leaves = F,
           show_trajectory_graph = F,
           min_expr = 1,
           alpha = 0.8,
           trajectory_graph_color = "grey28",
           ) +
  theme(
  axis.text = element_text(size = 6),    # Adjust the size as needed
  axis.title = element_text(size = 8),  # Adjust the size as needed
  legend.text = element_text(size = 4),  # Adjust the size as needed
  legend.title = element_text(size = 6),
  legend.key.size = unit(2, 'mm')) +
  scale_colour_viridis_c(option = "inferno")
dev.off()
###############################
GENES_HM<-c(Aortic_genes) #
pt.df<-as.data.frame(pt.matrix)
pt.df$gene_name<-rownames(pt.matrix)
ha = rowAnnotation(link = anno_mark(at = which(pt.df$gene_name %in% GENES_HM),
                   labels = as.character(pt.df[which(pt.df$gene_name %in% GENES_HM), "gene_name"]),
                   labels_gp = gpar(fontsize = 8),
                   padding = unit(1, "mm"))
                   )
heatmap_combination<-ComplexHeatmap::pheatmap(pt.matrix, scale="row",
                    cluster_cols = F,
                    cluster_rows = T,
                    cutree_rows = 4,
                    # cutree_cols = 3,
                     fontsize_col = 5,
                     color = myColor,
                    annotation_names_col = FALSE,
                    show_colnames = F,
                     show_rownames = F,
                     border_color = NA,
                    annotation_colors = NA,
                    right_annotation = ha,
                    annotation_col = Index,
                    border = TRUE)
pdf(file = "../2_Output/2_TAA/4_Trajectory/Trajectory_Heatmap.pdf", height = 8, width = 5)
heatmap_combination
dev.off()
```

# Integrated and harmonized Analysis (FFPE - TAA and ACTA2)

Once the individual snRNA-seq datasets were interrogated for quality and
normalized, they were fully integrated using standard protocol within
Seurat. To accomplish this, shared cell populations were matched across
datasets ("anchors") were used to correct for technical differences
between datasets (i.e. batch effects), as well as to perform comparative
sn-RNA-seq analysis across experimental conditions (ACTA2-mut vs. TAA
vs. CON).

Data from Chou et al. 2022 were incorporated to provide a more
generalizeable understanding of cellular heterogeneity at single-nuclear
resolution. Once the individual snRNA-seq datasets were interrogated for
quality and normalized, they were fully integrated using standard
protocol within Seurat. To accomplish this, shared cell populations were
matched across datasets ("anchors") were used to correct for technical
differences between datasets (i.e. batch effects), as well as to perform
comparative sn-RNA-seq analysis across experimental conditions.

```{r Integrate, eval=FALSE}
options(future.globals.maxSize= 891289600000000000)
library(Seurat)
ACTA2_scaled <- readRDS("../1_Input/ACTA2_annotated.rds")
ACTA2_scaled$Disease <- "TAA_ACTA2"
control_scaled <- readRDS("../1_Input/control_annotated.rds")
control_scaled$Disease <- "TAA_NG"
ifnb <- merge(ACTA2_scaled, y = control_scaled, project = "ACTA2", assay = "RNA")
ifnb <- NormalizeData(ifnb)
ifnb <- FindVariableFeatures(ifnb)
ifnb <- ScaleData(ifnb)
ifnb <- RunPCA(ifnb)
ifnb <- FindNeighbors(ifnb, dims = 1:30, reduction = "pca")
ifnb <- FindClusters(ifnb, resolution = 2, cluster.name = "unintegrated_clusters")
ifnb <- RunUMAP(ifnb, dims = 1:30, reduction = "pca", reduction.name = "umap.unintegrated")
pdf("../2_Output/3_Merged/2_Clustering/Unharmonized_UMAP.pdf", height = 5, width = 10)
DimPlot(ifnb, reduction = "umap.unintegrated", group.by = c("Disease", "cell_type"))
dev.off()
# Harmonize
ifnb <- IntegrateLayers(object = ifnb, method = CCAIntegration, orig.reduction = "pca", new.reduction = "integrated.cca",
    verbose = FALSE)
ifnb[["RNA"]] <- JoinLayers(ifnb[["RNA"]]) # re-join layers after integration
ifnb <- FindNeighbors(ifnb, reduction = "integrated.cca", dims = 1:30)
ifnb <- FindClusters(ifnb, resolution = c(0.1,0.2,0.3,0.4,0.5))
ifnb <- RunUMAP(ifnb, dims = 1:30, reduction = "integrated.cca")
pdf("../2_Output/3_Merged/2_Clustering/Harmonized_UMAP.pdf", height = 4, width = 12)
DimPlot(ifnb, reduction = "umap", group.by = c("Disease", "cell_type", "RNA_snn_res.0.4"))
dev.off()
saveRDS(ifnb, "../1_Input/FFPE_integrated.rds")
```

### Cell Annotation

```{r merged.annotation, eval=FALSE}
library(Seurat)
library(ggrepel)
library(RColorBrewer)
TAA_integrated<-readRDS(file = "../1_Input/FFPE_integrated.rds")
TAA_integrated$Disease <- factor(TAA_integrated$Disease, levels = c("TAA_NG", "TAA_ACTA2"))
colors = c("coral2",
           "coral3",
           "azure4", 
           "steelblue4",
           "wheat",
           "cadetblue3",
           "gray",
           "azure2",
           "cadetblue",
           "dodgerblue2",
           "black")
TAA_integrated[["RNA"]] <- JoinLayers(TAA_integrated[["RNA"]])
set2_colors <- brewer.pal(n = 8, name = "Set2")
UMAP_0.1<-DimPlot(TAA_integrated, reduction = "umap", label = T, repel = T,label.size = 4, group.by = "RNA_snn_res.0.1", split.by = "orig.ident", cols = ) + NoLegend()
UMAP_0.2<-DimPlot(TAA_integrated, reduction = "umap", label = T, repel = T,label.size = 4, group.by = "RNA_snn_res.0.2", split.by = "orig.ident", cols = colors) + NoLegend()
UMAP_0.3<-DimPlot(TAA_integrated, reduction = "umap", label = T, repel = T,label.size = 4, group.by = "RNA_snn_res.0.3", split.by = "orig.ident", cols = colors) + NoLegend()
UMAP_0.4<-DimPlot(TAA_integrated, reduction = "umap", label = T, repel = T,label.size = 4, group.by = "RNA_snn_res.0.4", split.by = "orig.ident", cols = colors) + NoLegend()
UMAP_0.5<-DimPlot(TAA_integrated, reduction = "umap", label = T, repel = T,label.size = 4, group.by = "RNA_snn_res.0.5", split.by = "orig.ident", cols = colors) + NoLegend()
pdf(file = "../2_Output/3_Merged/2_Clustering/UMAP.pdf", height = 5, width = 4)
UMAP_0.1
UMAP_0.2
UMAP_0.3
UMAP_0.4
UMAP_0.5
dev.off()
pdf(file = "../2_Output/3_Merged/2_Clustering/UMAP_pub.pdf")
DimPlot(TAA_integrated, reduction = "umap", label = T, repel = T,label.size = 4, group.by = "RNA_snn_res.0.4", split.by = "orig.ident", cols = colors) + NoLegend()
dev.off()
#Identify Clusters corresponding with known gene markers:
VSMC1_genes<-c("PKD1", "COL6A2", "PDLIM7", "FLNA", "SMTN")
VSMC1_available <- intersect(VSMC1_genes, rownames(TAA_integrated))
VSMC2_genes<-c("MYH11", "ACTA2", "ITGA8", "PRKG1", "CRISPLD1")
VSMC2_available <- intersect(VSMC2_genes, rownames(TAA_integrated))
Fibroblast1_genes<-c("ABCA10", "C3", "ADGRD1", "FBLN1", "DCN")
Fibroblast1_available <- intersect(Fibroblast1_genes, rownames(TAA_integrated))
Fibroblast2_genes<-c("NFASC",  "SAMD5", "PRSS23","UACA","TEX41") #
Fibroblast2_available <- intersect(Fibroblast2_genes, rownames(TAA_integrated))
Fibromyocyte_genes<-c("ADAMTS1", "RGS6", "TNC") # , "ANGPT2", "DGKG", "GRIP2"
Fibromyocyte_available <- intersect(Fibromyocyte_genes, rownames(TAA_integrated))
EC1_genes<-c("DIPK2B", "ARHGEF15", "STC1", "FLT1") # , "NOTCH4"
EC1_available <- intersect(EC1_genes, rownames(TAA_integrated))
EC2_genes<-c("VWF", "BMPER", "BMX", "NOS1")
EC2_available <- intersect(EC2_genes, rownames(TAA_integrated))
NKT_genes<-c("SKAP1", "RIPOR2", "ITGAL", "CD96") #  "RBPJ", "FYN",
NKT_available <- intersect(NKT_genes, rownames(TAA_integrated))
Macrophage_genes<-c("MRC1", "LGMN", "F13A1", "RBM47") #
Macrophage_available <- intersect(Macrophage_genes, rownames(TAA_integrated))
Dendritic_genes<-c("ITGAX", "S100A9", "CSF3R", "CXCL8") #
Dendritic_available <- intersect(Dendritic_genes, rownames(TAA_integrated))
B_GENES<-c("IGHM", "IGHD") #
B_GENES <- intersect(B_GENES, rownames(TAA_integrated))
T_GENES<-c("CD3","CD200","CCR7") #
T_GENES <- intersect(T_GENES, rownames(TAA_integrated))

# Plot density function
library(Nebulosa)
# VSMC1_density<-plot_density(TAA_integrated, VSMC1_available, joint = TRUE, combine = FALSE, pal = "magma")
VSMC1_density<-plot_density(TAA_integrated, VSMC1_available, joint = TRUE, combine = FALSE, pal = "magma")
VSMC2_density<-plot_density(TAA_integrated, VSMC2_available, joint = TRUE, combine = FALSE, pal = "magma")
Fibroblast1_density<-plot_density(TAA_integrated, Fibroblast1_available, joint = TRUE, combine = FALSE, pal = "magma")
Fibroblast2_density<-plot_density(TAA_integrated, Fibroblast2_available, joint = TRUE, combine = FALSE, pal = "magma")
Fibromyocyte_density<-plot_density(TAA_integrated, Fibromyocyte_available, joint = TRUE, combine = FALSE, pal = "magma")
EC1_density<-plot_density(TAA_integrated, EC1_available, joint = TRUE, combine = FALSE, pal = "magma")
EC2_density<-plot_density(TAA_integrated, EC2_available, joint = TRUE, combine = FALSE, pal = "magma")
Macrophage_density<-plot_density(TAA_integrated, Macrophage_available, joint = TRUE, combine = FALSE, pal = "magma")
NKT_density<-plot_density(TAA_integrated, NKT_available, joint = TRUE, combine = FALSE, pal = "magma")
Dendritic_density<-plot_density(TAA_integrated, Dendritic_available, joint = TRUE, combine = FALSE, pal = "magma")
B_density<-plot_density(TAA_integrated, B_GENES, joint = TRUE, combine = FALSE, pal = "magma")
# T_density<-plot_density(TAA_integrated, T_GENES, joint = TRUE, combine = FALSE, pal = "magma")
pdf(file = "../2_Output/3_Merged/2_Clustering/FFPE_Density_plots.pdf", height = 3, width = 3)
# VSMC1_density
VSMC1_density
VSMC2_density
Fibroblast1_density
Fibroblast2_density
EC1_density
EC2_density
Macrophage_density
NKT_density
Dendritic_density
Fibromyocyte_density
B_density
T_density
dev.off()
# Overlay these gene markers onto the UMAP to identify clusters
pdf(file = "../2_Output/3_Merged/2_Clustering/FFPE_FeaturePlots.pdf")
FeaturePlot(TAA_integrated, reduction = "umap", label = T, features = Fibroblast1_genes)
FeaturePlot(TAA_integrated, reduction = "umap", label = T, features = Fibroblast2_genes)
FeaturePlot(TAA_integrated, reduction = "umap", label = T, features = EC1_genes)
FeaturePlot(TAA_integrated, reduction = "umap", label = T, features = EC2_genes)
FeaturePlot(TAA_integrated, reduction = "umap", label = T, features = NKT_genes)
FeaturePlot(TAA_integrated, reduction = "umap", label = T, features = Macrophage_genes)
# FeaturePlot(TAA_integrated, reduction = "umap", label = T, features = VSMC1_genes)
FeaturePlot(TAA_integrated, reduction = "umap", label = T, features = VSMC2_genes)
FeaturePlot(TAA_integrated, reduction = "umap", label = T, features = Dendritic_genes)
dev.off()
#Create a figure of cell-type specific markers overlying the UMAP
pdf(file = "../2_Output/3_Merged/2_Clustering/CellType_Differentiation.pdf")
plot_density(TAA_integrated, c("ACTA2", "FBLN1", "F13A1", "VWF", "STC1", "NFASC", "ITGAL", "ITGAX"), joint = TRUE, combine = FALSE, pal = "magma")
dev.off()
# Create cluster labels based on the feature plot delineations of cellular phenotypes
Idents(TAA_integrated) <- TAA_integrated$RNA_snn_res.0.4
TAA_integrated <- RenameIdents(object = TAA_integrated, 
                          `0` = "VSMC_2", 
                          `1` = "VSMC_1", 
                          `2` = "Macrophage",
                          `3` = "Fibroblast",
                          `4` = "EC",
                          `5` = "Fibromyocyte",
                          `6` = "NKT",
                          `7` = "Dendritic",
                          `8` = "Fibromyocyte",
                          `9` = "B Lymph",
                          `10` = "T Lymph")
TAA_integrated$cell_type <- Idents(TAA_integrated) # redefine the cell-types based on the clustering of the integration, above
TAA_integrated$cell_type <- factor(TAA_integrated$cell_type, 
                                   levels = c(
                                    "VSMC_1", 
                                    "VSMC_2",
                                    "Fibromyocyte",
                                    "Fibroblast", 
                                    "EC",
                                    "Macrophage",
                                    "NKT",
                                    "Dendritic",
                                    "B Lymph",
                                    "T Lymph")
                                   )
Idents(TAA_integrated) <- TAA_integrated$cell_type
colors_cell = c("coral2",
           "coral3",
          "cadetblue3",
           "steelblue4",
           "wheat",
           "azure4", 
           "gray",
           "azure2",
           "dodgerblue2",
           "black")
# Create the UMAP of cell-type specific clusters
UMAP_Disease<-DimPlot(TAA_integrated,  reduction = "umap",label = F, group.by = "Disease",label.size = 4,
                        cols = c("steelblue4",
                                 "goldenrod2"))
UMAP_CellTypes<-DimPlot(TAA_integrated,  reduction = "umap",label = T, label.size = 4,
                        cols = colors_cell) + NoLegend()
pdf("../2_Output/3_Merged/2_Clustering/UMAP_Disease.pdf", height = 5, width = 5)
UMAP_Disease
dev.off()
pdf("../2_Output/3_Merged/2_Clustering/UMAP_CellTypes.pdf", height = 4, width = 4)
UMAP_CellTypes
dev.off()
# Differential Expression
DEGs_CellTypes<-FindAllMarkers(TAA_integrated)
write.csv(DEGs_CellTypes, "../2_Output/3_Merged/3_Differential.Expression/DEGs_Clusters.csv")
# Create a dot-bplot of the top 5 markers for each 
library(scCustomize)
paletteLength <- 100
myColor <- colorRampPalette(c("dodgerblue4", "white", "coral2"))(paletteLength)
top5_markers <- Extract_Top_Markers(marker_dataframe = DEGs_CellTypes, num_genes = 10, named_vector = FALSE,
    make_unique = TRUE)
pdf(file = "../2_Output/3_Merged/2_Clustering/Dot.plot_top5markers.pdf", height = 10, width = 7)
Clustered_DotPlot(seurat_object = TAA_integrated, features = top5_markers, k = 10, colors_use_exp = myColor, assay = "RNA")
dev.off()
# Heatmap of Clusters
library(dplyr)
DEGs_CellTypes %>%
    group_by(cluster) %>%
    top_n(n = 50, wt = avg_log2FC) -> top10
write.csv(top10, "../2_Output/3_Merged/3_Differential.Expression/top10.csv")
pdf(file = "../2_Output/3_Merged/2_Clustering/Celltype_Heatmap.pdf", height = 10, width = 15)
DoHeatmap(TAA_integrated, features = top10$gene, size = 3, disp.min = -2, disp.max = 2) + scale_fill_gradientn(colors = c("dodgerblue4", "white", "coral2")) + labs(title = "Heatmap of Top-10 most variable genes within each cluster")
dev.off()
# Probability distribution of specific genes across all clusters/cell types
VlnPlot(TAA_integrated, features = c("VWF", "HDAC9", "ACTA2"))
DotPlot(TAA_integrated, features = VSMC1_genes) #+ RotatedAxis()
saveRDS(TAA_integrated, file = "../1_Input/TAA_snRNA.rds")
```

## Proportional Comparison

## UMAP and Cellular Proportions

```{r cluster_combined, results='hide', eval=FALSE}
# Cell-type Specific Differential Expression
library(ggplot2)
library(dplyr)
library(Seurat)
library(dittoSeq)
TAA.combined <- readRDS(file = "../1_Input/TAA_snRNA.rds")
#Figure 2C -  Proportional Graph
pdf(file = "../2_Output/3_Merged/2_Clustering/Proportional.Bar_Treatment.pdf", height = 4, width = 4)
dittoBarPlot(
  object = TAA_integrated,
  var = "cell_type",
  group.by = "Disease"
) + 
  ggtitle(NULL) +
  scale_fill_manual(values = 
    c(VSMC_1="coral2",
     VSMC_2="coral3",
     Fibroblast="steelblue4", 
     EC="wheat",
     Macrophage="azure4",
     NKT="gray",
     Dendritic="tan2",
     Fibromyocyte="cadetblue3",
     `B Lymph`="darkgray",
     `T Lymph`="black")
    )
dev.off()

BarPlot <- dittoBarPlot(
    object = TAA.combined,
    var = "cell_type",
    group.by = "Disease", 
    data.out = T)
# Extract Data for export
BarPlot_sampledata <- BarPlot$data %>% 
  select(-count, -label.count.total.per.facet) %>% 
  tidyr::pivot_wider(., names_from = "grouping", values_from = "percent")
openxlsx::write.xlsx(BarPlot_sampledata, "../2_Output/BarPlot_CellTypes.xlsx")
# Visualize the features/genes
VizDimLoadings(TAA.combined, dims = 1:5, reduction = "pca")
pdf(file = "../2_Output/PC_Heatmaps.pdf")
DimHeatmap(TAA.combined, dims = 1:15, cells = 500, balanced = TRUE)
dev.off()
cell_type_disease_counts <- TAA.combined@meta.data %>%
  group_by(Disease, cell_type = Idents(TAA.combined)) %>%
  summarise(cell_count = n())
write.csv(cell_type_disease_counts, "cell_types_Disease.csv")
```

# Differential Analysis

```{r DiffEx_VSMC, eval=FALSE}
library(Seurat)
library(dplyr)
library(ggtrace)
library(ggplot2)
library(ggrepel)
TAA.combined<-readRDS(file = "../1_Input/TAA_snRNA.rds")
Idents(TAA.combined) <- "Disease"

#Volcano Plot
# Load packages
library(dplyr)
library(ggplot2)
library(ggrepel)
library(openxlsx)
library(Nebulosa)
# Read data from the web
options(ggrepel.max.overlaps = Inf)
cell_types <- unique(TAA.combined$cell_type)
# Loop through each cell type
for (CELL in cell_types) {
  # Subset the Seurat object based on the cell type
  subset_seurat <- subset(TAA.combined, subset = cell_type == CELL)
  # Generate a UMAP plot for the subsetted cell type
  DimPlot(subset_seurat, reduction = "umap", label = TRUE, repel = TRUE, label.size = 4) + 
    NoLegend()
  # Differential Expression Analysis for the current cell type
  de_markers <- FindMarkers(subset_seurat, ident.1 = "TAA_ACTA2", ident.2 = "TAA_NG", min.pct = 0.25)
  # Save the results to an Excel file
  output_file <- paste0("../2_Output/3_Merged/3_Differential.Expression/", CELL, "_DEGs.xlsx")
  openxlsx::write.xlsx(de_markers, output_file, rowNames = TRUE)
    ##############
  # Violin Plots (top 5)
  ###############
  top_degs.up <- de_markers %>% filter(pct.1>0.5 | pct.2>0.5) %>% top_n(., n = 5, avg_log2FC)
  top_degs.down <- de_markers %>% filter(pct.1>0.5 | pct.2>0.5) %>% top_n(., n = 5, -avg_log2FC)
  plots_up <- VlnPlot(subset_seurat, features = rownames(top_degs.up), split.by = "Disease",
      pt.size = 0, combine = FALSE)
  plots_up <- lapply(seq_along(plots_up), function(i) {
    plot <- plots_up[[i]] + theme(axis.title.x = element_blank())  # Remove x-axis label
    if (i == length(plots_up)) {
      return(plot)  # Keep the legend in the last plot
    } else {
      return(plot + NoLegend())  # Remove legend in other plots
    }
  })
  plots_down <- VlnPlot(subset_seurat, features = rownames(top_degs.down), split.by = "Disease",
      pt.size = 0, combine = FALSE)
  plots_down <- lapply(seq_along(plots_down), function(i) {
    plot <- plots_down[[i]] + theme(axis.title.x = element_blank())  # Remove x-axis label
    if (i == length(plots_down)) {
      return(plot)  # Keep the legend in the last plot
    } else {
      return(plot + NoLegend())  # Remove legend in other plots
    }
  })
  pdf(paste0("../2_Output/3_Merged/3_Differential.Expression/", CELL, "_TAA_ACTA2.v.NG_UP.pdf"), height = 3, width = 10)
  print(wrap_plots(plots = plots_up, ncol = 5))
  dev.off()
  pdf(paste0("../2_Output/3_Merged/3_Differential.Expression/", CELL, "_TAA_ACTA2.v.NG_DOWN.pdf"), height = 3, width = 10)
  print(wrap_plots(plots = plots_down, ncol = 5))
  dev.off()
  # Prepare data for Volcano Plot
    max_pval <- min(de_markers$p_val[de_markers$p_val != 0])  # Calculate the highest non-zero p_val value
    results <- de_markers %>%
      mutate(p_val = ifelse(p_val == 0, max_pval, p_val)) %>%  # Replace p_val == 0 with max_pval
      mutate(minuslogpvalue = -log(p_val), log2FC = avg_log2FC) %>%
      filter(p_val != 0) %>%
      mutate(sig = ifelse(p_val < 10^-8 & log2FC > 2, "P < 10^-8 and Fold-Change > 2", ifelse(p_val < 10^-8 & log2FC < -2, "P < 10^-8 and Fold-Change < -2", "Not Sig"))) %>%
      mutate(gene_name = rownames(.))  # Add gene_name as a new column
  #############
  #### Volcano Plot (TAA_ACTA2 vs. TAA_NG)
  ############
p <- ggplot(results, aes(log2FC, minuslogpvalue)) + 
  theme_classic() +
  geom_point(aes(fill = sig, size = minuslogpvalue), colour = "black", shape = 21, stroke = 0, alpha = 7/10) +
  geom_segment(aes(x = 2, xend = 2, y = 0, yend = max(results$minuslogpvalue, na.rm = TRUE)), size = 0.5, linetype = "dashed", lineend = "round") +  # Vertical line at x = 2
  geom_segment(aes(x = -2, xend = -2, y = 0, yend = max(results$minuslogpvalue, na.rm = TRUE)), size = 0.5, linetype = "dashed", lineend = "round") +  # Vertical line at x = -2
  geom_segment(aes(x = -5, xend = 5, y = 0 - log(10^-8), yend = 0 - log(10^-8)), size = 0.5, linetype = "dashed", lineend = "round") +
  geom_segment(aes(x = -5, xend = 5, y = 0, yend = 0), size = 1, linetype = "solid", lineend = "round") +
  geom_segment(aes(x = 0, xend = 0, y = 0, yend = max(results$minuslogpvalue, na.rm = TRUE)), size = 1, linetype = "solid", lineend = "round") +  # Vertical line at x = 0
  labs(x = expression(Log[2](Fold-Change)), y = expression(-Log[10](P-value))) + 
  scale_fill_manual(values = c("darkgray", "dodgerblue4", "goldenrod2")) +
  scale_size_continuous(range = c(0.1, 3)) +
  theme(axis.title.x = element_text(color = "black", size = 12),  # X-axis label black and larger
        axis.title.y = element_text(color = "black", size = 12),  # Y-axis label black and larger
        axis.text = element_text(color = "black", size = 10),  # Hide axis text
        axis.ticks = element_blank(),  # Hide axis ticks
        legend.position = "none")  # Hide legend
  # Annotate significant genes
  p <- p + 
    geom_text_repel(data = top_n(filter(results, log2FC < -2), 10, minuslogpvalue), aes(label = gene_name)) +
    geom_text_repel(data = top_n(filter(results, log2FC > 2), 10, minuslogpvalue), aes(label = gene_name))
  # Save the Volcano Plot
  pdf_file <- paste0("../2_Output/3_Merged/3_Differential.Expression/", CELL, "_VolcanoPlot.pdf")
  pdf(pdf_file, height = 5, width = 8)
  print(p)
  dev.off()
  ##Enrichr
  #### Pathways
  library(enrichR)
  library(dplyr)
  DEGs_up <- de_markers %>% filter(avg_log2FC > 0) %>% top_n(500, abs(avg_log2FC))
  DEGs_down <- de_markers %>% filter(avg_log2FC < 0) %>% top_n(500, abs(avg_log2FC))  
  #############
  #### Wikipathways Enrichment
  ############
  dbs <- c("WikiPathway_2023_Human")
  enriched_up <- enrichr(rownames(DEGs_up), dbs)
  enrich_up<-enriched_up[[dbs]] %>% filter(Adjusted.P.value < 0.05)
  head(enrich_up)
  enriched_down <- enrichr(rownames(DEGs_down), dbs)
  enrich_down<-enriched_down[[dbs]] %>% filter(Adjusted.P.value < 0.05)
  head(enrich_down)
  library(openxlsx)
  file_path <- "../2_Output/3_Merged/3_Differential.Expression/Wikipathways_All.Celltypes_TAAng.vs.ACTA2.xlsx"
      if (file.exists(file_path)) { # Check if the file exists and load or create the workbook accordingly
        wb <- loadWorkbook(file_path)  # Load existing workbook
        message("Workbook loaded successfully.")
        existing_sheets <- getSheetNames(file_path)  # Get existing sheet names using getSheetNames() only if the file exists
      } else {
        wb <- createWorkbook()  # Create new workbook
        message("New workbook created.")
        existing_sheets <- character(0)  # Empty vector indicating no sheets exist
      }
      if (!(paste0(CELL, "_UP") %in% existing_sheets)) { # Adding sheets conditionally based on their existence
        addWorksheet(wb, sheetName = paste0(CELL, "_UP"))
        writeData(wb, sheet = paste0(CELL, "_UP"), enrich_up, startCol = 1)
      } else {
        message(paste("Sheet", paste0(CELL, "_UP"), "already exists. Skipping."))
      }
      if (!(paste0(CELL, "_DOWN") %in% existing_sheets)) {
        addWorksheet(wb, sheetName = paste0(CELL, "_DOWN"))
        writeData(wb, sheet = paste0(CELL, "_DOWN"), enrich_down, startCol = 1)
      } else {
        message(paste("Sheet", paste0(CELL, "_DOWN"), "already exists. Skipping."))
      }
      saveWorkbook(wb, file_path, overwrite = TRUE) # Save the workbook
      message("Workbook saved successfully.")
  #############
  #### GWAS Enrichment
  ############
  dbs <- c("GWAS_Catalog_2023")
  enriched_up <- enrichr(rownames(DEGs_up), dbs)
  enrich_up<-enriched_up[[dbs]] %>% filter(Adjusted.P.value < 0.05)
  head(enrich_down)
  enriched_down <- enrichr(rownames(DEGs_down), dbs)
  enrich_down<-enriched_down[[dbs]] %>% filter(Adjusted.P.value < 0.05)
  head(enrich_up)
  file_path <- "../2_Output/3_Merged/3_Differential.Expression/GWAS_All.Celltypes_TAAng.vs.ACTA2.xlsx"
      if (file.exists(file_path)) { # Check if the file exists and load or create the workbook accordingly
        wb <- loadWorkbook(file_path)  # Load existing workbook
        message("Workbook loaded successfully.")
        existing_sheets <- getSheetNames(file_path)  # Get existing sheet names using getSheetNames() only if the file exists
      } else {
        wb <- createWorkbook()  # Create new workbook
        message("New workbook created.")
        existing_sheets <- character(0)  # Empty vector indicating no sheets exist
      }
      if (!(paste0(CELL, "_UP") %in% existing_sheets)) { # Adding sheets conditionally based on their existence
        addWorksheet(wb, sheetName = paste0(CELL, "_UP"))
        writeData(wb, sheet = paste0(CELL, "_UP"), enrich_up, startCol = 1)
      } else {
        message(paste("Sheet", paste0(CELL, "_UP"), "already exists. Skipping."))
      }
      if (!(paste0(CELL, "_DOWN") %in% existing_sheets)) {
        addWorksheet(wb, sheetName = paste0(CELL, "_DOWN"))
        writeData(wb, sheet = paste0(CELL, "_DOWN"), enrich_down, startCol = 1)
      } else {
        message(paste("Sheet", paste0(CELL, "_DOWN"), "already exists. Skipping."))
      }
      saveWorkbook(wb, file_path, overwrite = TRUE) # Save the workbook
      message("Workbook saved successfully.")
}
```

# Comparing DEGs across Cell Types

```{r celltype_compare, eval=FALSE}
library(Seurat)
library(dplyr)
TAA.combined<-readRDS(file = "../1_Input/TAA_snRNA.rds")
cell_states <- unique(TAA.combined$cell_type)  # Get unique disease states
deg_list <- list()
# Loop through each disease state to find DEGs
for (CELL in cell_states) {
  markers <- openxlsx::read.xlsx(paste0("../2_Output/3_Merged/3_Differential.Expression/", CELL, "_DEGs.xlsx"), rowNames = T) %>% filter(p_val < 0.05)
  deg_list[[CELL]] <- rownames(markers[markers$p_val_adj < 0.05, ])  # Extract DEGs with p-value < 0.05
}
### Cell Type Comparison
# Create a data frame from the list for UpSetR
deg_matrix <- data.frame(
  gene = unique(unlist(deg_list)),  # Unique genes
  stringsAsFactors = FALSE
)
# Add binary presence/absence indicators for each cell type
for (cell_type in names(deg_list)) {
  deg_matrix[[cell_type]] <- ifelse(deg_matrix$gene %in% deg_list[[cell_type]], 1, 0)
}
desired_order <- rev(c("VSMC_1", "VSMC_2", "Fibromyocyte", "Fibroblast", "Macrophage", "Dendritic", "NKT", "EC"))
t <- upset(deg_matrix, 
      sets = desired_order,  # Names of the cell types
      # text.scale = c(2, 1.5, 2, 2, 2, 2),
      mb.ratio = c(0.5, 0.5),
      keep.order = TRUE,  # Keep the order of sets as provided
      main.bar.color = "#56B4E9",  # Color for main bars
      sets.bar.color = "#D55E00",  # Color for sets bars
      order.by = "freq")  # Order by frequency of intersections
t
pdf("../2_Output/3_Merged/3_Differential.Expression/Venn_UpSetR.pdf", height = 4, width = 7)
print(t)
dev.off()
# Load libraries
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggplot2)
library(ReactomePA)
for (CELL in cell_states) {
  markers <- openxlsx::read.xlsx(paste0("../2_Output/3_Merged/3_Differential.Expression/", CELL, "_DEGs.xlsx"), rowNames = T) %>% top_n(n = 100, wt = abs(avg_log2FC))
  deg_list[[CELL]] <- rownames(markers[markers$p_val_adj < 0.05, ])  # Extract DEGs with p-value < 0.05
}
# Prepare DEG list (Example DEG list structure)
deg_list <- deg_list[!names(deg_list) %in% c("B Lymph","T Lymph")]
common_genes <- Reduce(intersect, deg_list)
# Convert gene symbols to Entrez IDs
deg_list_entrez <- lapply(deg_list, function(genes) {
  bitr(genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)$ENTREZID
})
# Perform Reactome pathway enrichment analysis for each cell type
reactome_results <- lapply(deg_list_entrez, function(entrez_genes) {
  enrichPathway(gene = entrez_genes, organism = "human", qvalueCutoff = 0.05)
})
# Assign cell type names
names(reactome_results) <- names(deg_list)
# Create a compareCluster object for Reactome pathway enrichment
compare_cluster_results <- compareCluster(
  geneCluster = deg_list_entrez,
  fun = "enrichPathway",
  organism = "human",
  pvalueCutoff = 0.05
)
# compare_cluster_results <- compareCluster(
#   geneCluster = deg_list_entrez,  # List of genes for each cell type in Entrez ID format
#   fun = "enrichKEGG",             # Use "enrichKEGG" for KEGG pathway analysis
#   organism = "hsa",               # Specify organism as "hsa" for human in KEGG
#   pvalueCutoff = 0.05             # p-value cutoff for pathway enrichment
# )
# compare_cluster_results <- compareCluster(
#   geneCluster = deg_list_entrez,  # List of genes for each cell type in Entrez ID format
#   fun = "enrichDO",             # Use "enrichKEGG" for KEGG pathway analysis
#   organism = "hsa",               # Specify organism as "hsa" for human in KEGG
#   pvalueCutoff = 0.05             # p-value cutoff for pathway enrichment
# )
# compare_cluster_results <- compareCluster(
#   geneCluster = deg_list_entrez,
#   fun = "enrichWP",
#   organism = "Homo sapiens",
#   pvalueCutoff = 0.05
# )
# compare_cluster_results <- compareCluster(
#   geneCluster = deg_list_entrez,
#   fun = "enrichDGN",
#   pvalueCutoff = 0.01
# )

###############

compare_cluster_results@compareClusterResult$cell_type <- compare_cluster_results@compareClusterResult$Cluster
pdf("../2_Output/3_Merged/3_Differential.Expression/DotPlot_SPLIT.pdf", height = 4, width = 11)
dotplot(compare_cluster_results, showCategory = 3, x = "GeneRatio", split = "cell_type", label_format = 50) +
  facet_grid(. ~ cell_type) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 10))
dev.off()

# Visualize using dotplot
pdf("../2_Output/3_Merged/3_Differential.Expression/DotPlot.pdf", height = 4, width = 9)
dotplot(compare_cluster_results, color = "p.adjust", showCategory = 2) + 
  scale_x_discrete(labels = function(x) gsub("\\s*\\(\\d+\\)$", "", x)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        xis.text.y = element_text(size = 6))
dev.off()

pathway_genes <- compare_cluster_results@compareClusterResult %>%
  filter(Description == "Extracellular matrix organization") %>%
  pull(geneID) %>%
  strsplit("/") %>%
  unlist()
# Step 2: Convert Entrez IDs to gene symbols
converted_genes <- bitr(
  pathway_genes,
  fromType = "ENTREZID",
  toType = "SYMBOL",
  OrgDb = org.Hs.eg.db
)
gene_symbols <- converted_genes$SYMBOL
genes_curated <- c("TIMP1", "COL8A1", "EFEMP1", "COL14A1", "SERPINE1", "ADAMTS1", "VCAN", "ITGA7", "A2M", "ITGA8", "MMP2", "HSPG2")
 # Violin Plots (top 5)
  ###############
library(Seurat)
library(patchwork)
  plots_up <- VlnPlot(TAA.combined, features = common_genes, split.by = "Disease", group.by = "cell_type",
      pt.size = 0, combine = FALSE)
  plots_up <- lapply(seq_along(plots_up), function(i) {
    plot <- plots_up[[i]] + theme(axis.title.x = element_blank())  # Remove x-axis label
    if (i == length(plots_up)) {
      return(plot)  # Keep the legend in the last plot
    } else {
      return(plot + NoLegend())  # Remove legend in other plots
    }
  })
  pdf(paste0("../2_Output/3_Merged/3_Differential.Expression/Conserved.pdf"), height = 6, width = 10)
  print(wrap_plots(plots = plots_up, ncol = 3))
  dev.off()
```

# Cell-cell Interactions

```{r LIANA, eval=FALSE}
# Load necessary libraries
library(Seurat)
library(tidyverse)
library(magrittr)
library(liana)
library(openxlsx)
library(circlize)
library(ComplexHeatmap)
# Create folder structure for output if it does not already exist
ifelse(!dir.exists(file.path("../2_Output/Regulation/")), dir.create(file.path("../2_Output/Regulation/")), FALSE)

# Import the Seurat object containing the ACTA2-mutant sample
acta2_scaled <- readRDS(file = "../1_Input/ACTA2_annotated.rds")

# Subset data to include only relevant cell types if needed, e.g., VSMCs, ECs, Myeloid cells, etc.
# Uncomment and modify the line below if specific cell types need to be isolated
# acta2_scaled <- subset(acta2_scaled, idents = c("VSMC", "EC", "Myeloid", "Fibroblast"))

# Run LIANA to identify potential cell-cell interactions
liana_results <- liana_wrap(acta2_scaled) # Use the appropriate resource if working with human data
liana_results <- liana_aggregate(liana_results)

# Filter LIANA results for significant interactions (e.g., p-value <= 0.05)
liana_filtered <- liana_results %>% filter(cellphonedb.pvalue <= 0.05)

# Save filtered results to an Excel file
openxlsx::write.xlsx(liana_filtered, "../2_Output/Regulation/LIANA_ACTA2_Mutant_Significant.xlsx")

# Check unique targets
unique_targets <- unique(liana_results$target)
print(unique_targets)  # Print to ensure it's populated correctly

# Create a heatmap of interaction frequencies
pdf(file = "../2_Output/Regulation/LIANA_Heatmap_ACTA2_Mutant.pdf", height = 4, width = 5)
heat_freq(liana_filtered)
dev.off()

# Visualize cell-cell interactions using a dot plot
pdf(file = "../2_Output/Regulation/LIANA_DotPlot_ACTA2_Mutant.pdf", height = 7, width = 7)
liana_results %>%
  liana_dotplot(source_groups = c("VSMC_2"),  # Define the source cell type(s)
                target_groups = unique_targets[!is.na(unique_targets)],  # Remove NA if any
                ntop = 20) + 
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1), 
        plot.title = element_text(size = 0), 
        axis.title.x = element_text(size = 0))
dev.off()

# Define color scheme for each cell type, including all targets in unique_targets
grid.col <-c(VSMC_1="#b95e4b",
             VSMC_2="coral4",
             VSMC_3="coral3",
             Fibroblast="wheat", 
             EC1="steelblue4",
             EC2="deepskyblue3",
             Macrophage="azure4",
             NKT="gray",
             Dendritic="tan2")

# Aggregate interactions to create a weight for each source-target pair
liana_data_aggregated <- liana_filtered %>%
  mutate(weight = -log10(aggregate_rank)) %>%  # Convert rank to weight for visualization
  group_by(source, target) %>%
  summarise(weight = sum(weight, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(!is.infinite(weight) & weight > 0)  # Exclude infinite or zero weights

# Convert to data frame for chord diagram compatibility
liana_data_aggregated <- as.data.frame(liana_data_aggregated)

# Save the chord diagram to a PDF
pdf(file = "../2_Output/Regulation/LIANA_ChordDiagram_ACTA2_Mutant.pdf", height = 6, width = 9)
circlize::chordDiagram(
  x = liana_data_aggregated,
  grid.col = grid.col,
  transparency = 0.5,
  directional = 1,
  direction.type = c("diffHeight", "arrows"),
  link.arr.type = "big.arrow",
  annotationTrack = "grid",
  preAllocateTracks = list(track.height = 0.1)
)

# Add cell type labels with custom spacing
circlize::circos.trackPlotRegion(track.index = 1, panel.fun = function(x, y) {
  circos.text(CELL_META$xcenter, CELL_META$ylim[1] + 0.3, CELL_META$sector.index, facing = "bending", niceFacing = TRUE, adj = c(0.5, 0.5))
}, bg.border = NA)

# Create and draw a legend for cell types
legend <- ComplexHeatmap::Legend(
  at = names(grid.col), 
  title = "Cell Type", 
  legend_gp = gpar(fill = grid.col)
)
ComplexHeatmap::draw(legend, x = unit(1, "npc") - unit(5, "mm"), just = "right")

# Close the PDF device
dev.off()
library(Nebulosa)
plot_density(acta2_scaled, "MYL9", joint = TRUE, combine = FALSE, pal = "magma")
FeaturePlot(acta2_scaled, features = "LGALS1", reduction = "umap") +
    scale_color_viridis_c(option = "magma") +  # Apply the "mma" color scale
    ggtitle("UMAP Plot Showing Expression of LGALS1") +
    theme_minimal()  # Optional: Apply a minimal theme for a cleaner appearance

```

# ShinyR App Development

```{r shiny, eval=FALSE}
library(Seurat)
library(ShinyCell)
TAA.combined<-readRDS(file = "../1_Input/TAA_snRNA.rds")
library(clusterProfiler)
library(org.Mm.eg.db)  # Mouse gene database
# Define the mouse GO term of interest, e.g., GO:0008150 (biological process)
go_term <- "GO:0008283" # used in the original 2023 paper
# Extract gene symbols associated with the specified GO term
genes_vector <- AnnotationDbi::select(
  org.Mm.eg.db,                 # Mouse gene database
  keys = go_term,               # GO term of interest
  columns = "SYMBOL",           # Retrieve the gene symbols
  keytype = "GOALL"             # Define the key type as GO terms
)$SYMBOL
# Proliferation Score
genes_vector <- toupper(genes_vector)
gene_list <- genes_vector[genes_vector %in% rownames(TAA.combined)] 
gene_expr <- GetAssayData(object = TAA.combined, slot = "data")[gene_list, ]
sum_expr <- colSums(gene_expr)
TAA.combined$Proliferation <- scale(sum_expr)
# Load required libraries
library(Seurat)
library(ShinyCell)

# Make a config file
scConf <- createConfig(
  obj = TAA.combined,                # Your Seurat object
  meta.to.include = c("cell_type", "Disease", "Proliferation", "RNA_snn_res.0.4"),  # Metadata columns to include
  legendCols = 4,                    # Number of columns in categorical metadata legends
  maxLevels = 50                     # Max number of levels for categorical metadata
)
makeShinyApp(
  obj = TAA.combined,               # Seurat object or file path to the dataset
  scConf = scConf,                  # Configuration data.table generated in Step 1
  default.dimred = c("umap_1", "umap_2"),
  shiny.footnotes = "Pepin et al. 2024",
  gex.assay = "RNA",                # Assay in Seurat object to use (e.g., "RNA", "integrated")
  gex.slot = "data",                # Slot in the Seurat assay to use ("data" is default for normalized expression)
  shiny.title = "ACTA2-p.Met49Thr: snRNA-Sequencing Analysis",  # Title for the Shiny app
  shiny.dir = "ShinyCellApp",       # Directory to create the app files
  enableSubset = TRUE,              # Enable subsetting cells functionality in the app
  defPtSiz = 1.25,                  # Default point size for single cells in plots
  default.gene1 = "MYH11",          # Primary default gene to show in feature plots
  default.gene2 = "KLF4"            # Secondary default gene for comparison
)
```

# Supplemental Table: R Session Information

All packages and setting are acquired using the following command.

```{r settings}
# # unloading system variables after session
# homer_tE<-Sys.time()
# homer_t<-homer_tE - homer_tS
# homer_t
# 
# end_t<-Sys.time()
# Total_time<-end_t - start_t
# Total_time
# Sys.unsetenv("motifLoc_UP")
# Sys.unsetenv("motifLoc_DOWN")
# Sys.unsetenv("motifLoc_WD")
# Write 
options(kableExtra.latex.load_packages = FALSE)
library(kableExtra)
sinfo<-devtools::session_info()
sinfo$platform
sinfo$packages %>% kable(
                         align="c",
                         longtable=T,
                         booktabs=T,
                         caption="Packages and Required Dependencies") %>%
    kable_styling(latex_options=c("striped", "repeat_header", "condensed"))
```
